
### Calculating exposome multivariable models

```{r packages, eval=TRUE}
library(Hmisc)
library(survival)
library(mice)
library(CoxR2)
library(car)
library(DescTools)
library(ggplot2)
library(reshape2)
library(rms)
library(viridis)

```

```{r load, results='hide'}

path <- "/path/to/file"

# load analysis datasets
load(paste0(path, "/ACM_final_analysis_datasets_aug_20_2022_all_data.RData"))
load(paste0(path, "/ACM_final_analysis_datasets_aug_20_2022_scotwales.RData"))

# load physical environment PCA results
load(paste0(path, "/physical_environment_PCA_nov_07_2022_all_sexes.RData"))

# load NCD data
load("/path/to/file/NCD_all_subsets_aug_04_2022.RData")

# load interview baseline dx data
load("/path/to/file/baseline_dx_data_july_07_2022.RData")

# load variable categories
categories <- read.csv("/path/to/file/Argentieri UKB dataset data dictionary.csv")

# load PRS data
load(paste0(path, "/datasets/PRS_data_nov_04_2022.RData"))

# load disease analysis results
load(paste0(path, "/all_disease_analysis_output_with_cs_nov_07_2022_all_sexes.RData"))

# load list of vars significant in cluster multivariate analyses
exposures <- read.csv(paste0(path, "/total_sig_vars_list_nov_07_2022_all_sexes.csv"))
exposures <- as.vector(unlist(exposures[1]))

# set global R options
options(scipen = 999) # turn off scientific notation
pboptions(type = "timer", char = "=") # initialize progress bar
cores <- NULL
```

## Data prep

```{r variable_prep}

# make new var for scaled age
for (k in seq_along(all_data)) {
    all_data[[k]]$age_scaled <- all_data[[k]]$recruitment_age
    scotwales[[k]]$age_scaled <- scotwales[[k]]$recruitment_age
}

# scale function
scale_numeric <- function(x) { 
    # create logical vector of numeric columns
    nums <- sapply(x, is.numeric)
    
    # list of numeric variables to exclude from scaling
    scale_exclude <- c(
        Cs(
            censor_age,
            recruitment_age,
            ACM_event_indicator,
            eid,
            time,
            ACM_survival_time,
            hazard     
        )
    )
    
    # exclude those columns I don't want to scale
    nums[names(nums) %in% scale_exclude] <- FALSE # don't want to scale these
    
    # # don't bother scaling other numeric cols not in analysis
    nums[names(nums) %nin% exposures] <- FALSE

    # perform scale
    x[nums] <- scale(x[nums],
                     scale = TRUE)
    
    # return entire dataset
    return(x)
}

# scale desired numeric columns
all_data <- lapply(all_data, scale_numeric)
scotwales <- lapply(scotwales, scale_numeric)

```

## Run physical environment PCAs in Scotland/Wales datasets

```{r pca_scotwales, eval=FALSE}

# duplicate dataset for PCA
data_pca <- scotwales

# make rownames eid
for (k in seq_along(data_pca)) {
    rownames(data_pca[[k]]) <- data_pca[[k]]$eid
}

# all pollution vars significant in XWAS
pollution_vars <- 
    c(
        Cs(
            NO_2010,
            NO2_2005,
            NO2_2006,
            NO2_2007,
            NO2_2010,
            PM10_2007,
            PM10_2010,
            PM2.5_2010,
            PM2.5_absorbance_2010
        )
    )

# all greenspace and natural environment vars significant in XWAS
greenspace_vars <- 
    c(
        Cs(
            greenspace_buffer_1000m,
            greenspace_buffer_300m,
            natural_environment_buffer_1000m,
            natural_environment_buffer_300m
        )
    )

# run PCA in each imputed dataset - set scale to T because not scaled above
set.seed(3456)
pollution_PCA_list <- 
    lapply(data_pca, function(x) {
        prcomp(x[which(names(x) %in% pollution_vars)], 
               scale = TRUE)
    }
)

set.seed(3456)
greenspace_PCA_list <- 
    lapply(data_pca, function(x) {
        prcomp(x[which(names(x) %in% greenspace_vars)], 
               scale = TRUE)
    }
)

# summary to see explained variance per PC
summary(pollution_PCA_list[[1]])
summary(pollution_PCA_list[[2]])
summary(pollution_PCA_list[[3]])
summary(pollution_PCA_list[[4]])
summary(pollution_PCA_list[[5]])
# first 3 PCs explain > 98% of variance in all datasets

summary(greenspace_PCA_list[[1]])
summary(greenspace_PCA_list[[2]])
summary(greenspace_PCA_list[[3]])
summary(greenspace_PCA_list[[4]])
summary(greenspace_PCA_list[[5]])
# first 2 PCs explain > 96% of variance in all datasets

# get df of pollution PC scores for each participant
pollution_pca_dfs_scotwales <- 
    lapply(pollution_PCA_list, function(x) {
        # get individual PC scores
        df <- data.frame(x$x)
        # subset to just 3 PCs
        df <- df[1:4]
        # rename columns
        colnames(df) <- c("air_pollution_PC1", 
                          "air_pollution_PC2",
                          "air_pollution_PC3",
                          "air_pollution_PC4")
        # return df
        return(df)
    }
)


# get df of greenspace PC scores for each participant
greenspace_PCA_dfs_scotwales <- 
    lapply(greenspace_PCA_list, function(x) {
        # get individual PC scores
        df <- data.frame(x$x)
        # subset to just 2 PCs
        df <- df[1:2]
        # rename columns
        colnames(df) <- c("greenspace_PC1", 
                          "greenspace_PC2")
        # return df
        return(df)
    }
)


# make eids a column again for merge with larger dataset
for (k in seq_along(pollution_pca_dfs_scotwales)) {
    pollution_pca_dfs_scotwales[[k]]$eid <- rownames(pollution_pca_dfs_scotwales[[k]])
}

# make eids a column again for merge with larger dataset
for (k in seq_along(greenspace_PCA_dfs_scotwales)) {
    greenspace_PCA_dfs_scotwales[[k]]$eid <- rownames(greenspace_PCA_dfs_scotwales[[k]])
}

## make plots for PCA loadings

pca1 <- pollution_PCA_list[[1]]
pca2 <- greenspace_PCA_list[[1]]

pca_mat1 <- as.matrix(pca1$rotation[,1:3])
pca_mat1 <- ifelse(pca_mat1 > -0.3 & pca_mat1 < 0.3, 
                  0,
                  pca_mat1)
melted_cormat1 <- melt(pca_mat1)

pca_mat2 <- as.matrix(pca2$rotation[,1:2])
pca_mat2 <- ifelse(pca_mat2 > -0.3 & pca_mat2 < 0.3, 
                  0,
                  pca_mat2)
melted_cormat2 <- melt(pca_mat2)

heat1 <- 
    ggplot(data = melted_cormat1, aes(x=Var2, y=Var1, fill=value)) + 
    geom_tile(color = "white") +
    scale_fill_gradient2(
        low = "blue",
        high = "red",
        mid = "white",
        midpoint = 0,
        limit = c(-1, 1),
        space = "Lab",
        name = ""
    ) +
    ggtitle("PCA loadings for physical environment variables - Scottish/Welsh UK Biobank participants") +
    theme_classic(base_size = 16) +   
    theme(
        legend.position = "right",
        legend.title = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(
            size = 14,
            face = "italic",
            margin = margin(0,0,30,0)
        )
    ) +
    labs(x = "",
         y = "",
         subtitle = "Note: only loadings < -0.3 or > 0.3 are colored"
    ) 

heat2 <- 
    ggplot(data = melted_cormat2, aes(x=Var2, y=Var1, fill=value)) + 
    geom_tile(color = "white") +
    scale_fill_gradient2(
        low = "blue",
        high = "red",
        mid = "white",
        midpoint = 0,
        limit = c(-1, 1),
        space = "Lab",
        name = ""
    ) +
    # ggtitle("PCA loadings for natural/green space variables - UK Biobank") +
    theme_classic(base_size = 16) +   
    theme(
        legend.position = "right",
        legend.title = element_blank(),
    ) +
    labs(x = "",
         y = ""
    ) 

heat1
heat2

library(gridExtra)
grid <- grid.arrange(heat1, heat2, ncol = 1, nrow = 2, widths = c(6), heights = c(10, 5))

ggsave(grid, filename = paste0(path, "/phys_environment_PCA_sept_10_2022_scotwales.png"), width = 9, height = 6)


# merge PCA dfs together and save
pca_dfs_scotwales <- as.list(seq_along(greenspace_PCA_dfs_scotwales))
for (k in seq_along(pca_dfs_scotwales)) {
    pca_dfs_scotwales[[k]] <-
        merge(pollution_pca_dfs_scotwales[[k]],
              greenspace_PCA_dfs_scotwales[[k]],
              by = "eid",
              sort = FALSE)
}

# according to loadings, PC1 is inversely related to the pollution vars. So flip sign for sensible interpretation.
for (k in seq_along(pca_dfs_scotwales)) {
    pca_dfs_scotwales[[k]]$air_pollution_PC1 <- pca_dfs_scotwales[[k]]$air_pollution_PC1 * (-1)
}

save(pca_dfs_scotwales, file = paste0(path, "/results/physical_environment_PCA_sept_10_2022_scotwales.RData"))

# scale for analyses
for (k in seq_along(pca_dfs_scotwales)) {
pca_dfs_scotwales[[k]][-1] <- scale(pca_dfs_scotwales[[k]][-1], scale = TRUE)
}

```

## PRS data processing

```{r PRS_quintiles}

# breast cancer quintiles
PRS_data$PRS_breast_quintile <- NA
PRS_data$PRS_breast_quintile <- cut(PRS_data$PRS_breast, quantile(PRS_data$PRS_breast, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_breast_quintile)
PRS_data$PRS_breast_quintile <- factor(PRS_data$PRS_breast_quintile, ordered = FALSE)
PRS_data$PRS_breast_quintile <- relevel(PRS_data$PRS_breast_quintile, ref = "(-4.98,-0.977]")

# bowel cancer quintiles
PRS_data$PRS_bowel_quintile <- NA
PRS_data$PRS_bowel_quintile <- cut(PRS_data$PRS_bowel, quantile(PRS_data$PRS_bowel, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_bowel_quintile)
PRS_data$PRS_bowel_quintile <- factor(PRS_data$PRS_bowel_quintile, ordered = FALSE)
PRS_data$PRS_bowel_quintile <- relevel(PRS_data$PRS_bowel_quintile, ref = "(-4.67,-0.664]")

# prostate cancer quintiles
PRS_data$PRS_prostate_quintile <- NA
PRS_data$PRS_prostate_quintile <- cut(PRS_data$PRS_prostate, quantile(PRS_data$PRS_prostate, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_prostate_quintile)
PRS_data$PRS_prostate_quintile <- factor(PRS_data$PRS_prostate_quintile, ordered = FALSE)
PRS_data$PRS_prostate_quintile <- relevel(PRS_data$PRS_prostate_quintile, ref = "(-4.44,-0.271]")

# T2D quintiles
PRS_data$PRS_T2D_quintile <- NA
PRS_data$PRS_T2D_quintile <- cut(PRS_data$PRS_T2D, quantile(PRS_data$PRS_T2D, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_T2D_quintile)
PRS_data$PRS_T2D_quintile <- factor(PRS_data$PRS_T2D_quintile, ordered = FALSE)
PRS_data$PRS_T2D_quintile <- relevel(PRS_data$PRS_T2D_quintile, ref = "(-4.57,-0.945]")

# CVD quintiles
PRS_data$PRS_CVD_quintile <- NA
PRS_data$PRS_CVD_quintile <- cut(PRS_data$PRS_CVD, quantile(PRS_data$PRS_CVD, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_CVD_quintile)
PRS_data$PRS_CVD_quintile <- factor(PRS_data$PRS_CVD_quintile, ordered = FALSE)
PRS_data$PRS_CVD_quintile <- relevel(PRS_data$PRS_CVD_quintile, ref = "(-4.8,-0.928]")

# CAD quintiles
PRS_data$PRS_CAD_quintile <- NA
PRS_data$PRS_CAD_quintile <- cut(PRS_data$PRS_CAD, quantile(PRS_data$PRS_CAD, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_CAD_quintile)
PRS_data$PRS_CAD_quintile <- factor(PRS_data$PRS_CAD_quintile, ordered = FALSE)
PRS_data$PRS_CAD_quintile <- relevel(PRS_data$PRS_CAD_quintile, ref = "(-4.95,-0.968]")

# stroke quintiles
PRS_data$PRS_stroke_quintile <- NA
PRS_data$PRS_stroke_quintile <- cut(PRS_data$PRS_stroke, quantile(PRS_data$PRS_stroke, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_stroke_quintile)
PRS_data$PRS_stroke_quintile <- factor(PRS_data$PRS_stroke_quintile, ordered = FALSE)
PRS_data$PRS_stroke_quintile <- relevel(PRS_data$PRS_stroke_quintile, ref = "(-4.22,-0.805]")

# Alzheimer's quintiles
PRS_data$PRS_alzheimers_quintile <- NA
PRS_data$PRS_alzheimers_quintile <- cut(PRS_data$PRS_alzheimers, quantile(PRS_data$PRS_alzheimers, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_alzheimers_quintile)
PRS_data$PRS_alzheimers_quintile <- factor(PRS_data$PRS_alzheimers_quintile, ordered = FALSE)
PRS_data$PRS_alzheimers_quintile <- relevel(PRS_data$PRS_alzheimers_quintile, ref = "(-3.73,-0.776]")

# parkinson's quintiles
PRS_data$PRS_parkinsons_quintile <- NA
PRS_data$PRS_parkinsons_quintile <- cut(PRS_data$PRS_parkinsons, quantile(PRS_data$PRS_parkinsons, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_parkinsons_quintile)
PRS_data$PRS_parkinsons_quintile <- factor(PRS_data$PRS_parkinsons_quintile, ordered = FALSE)
PRS_data$PRS_parkinsons_quintile <- relevel(PRS_data$PRS_parkinsons_quintile, ref = "(-4.55,-0.998]")

# rheumatoid quintiles
PRS_data$PRS_rheumatoid_quintile <- NA
PRS_data$PRS_rheumatoid_quintile <- cut(PRS_data$PRS_rheumatoid, quantile(PRS_data$PRS_rheumatoid, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_rheumatoid_quintile)
PRS_data$PRS_rheumatoid_quintile <- factor(PRS_data$PRS_rheumatoid_quintile, ordered = FALSE)
PRS_data$PRS_rheumatoid_quintile <- relevel(PRS_data$PRS_rheumatoid_quintile, ref = "(-3.74,-0.709]")

# macular degeneration quintiles
PRS_data$PRS_macular_quintile <- NA
PRS_data$PRS_macular_quintile <- cut(PRS_data$PRS_macular, quantile(PRS_data$PRS_macular, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_macular_quintile)
PRS_data$PRS_macular_quintile <- factor(PRS_data$PRS_macular_quintile, ordered = FALSE)
PRS_data$PRS_macular_quintile <- relevel(PRS_data$PRS_macular_quintile, ref = "(-4.32,-0.754]")

# osteoporosis quintiles
PRS_data$PRS_osteoporosis_quintile <- NA
PRS_data$PRS_osteoporosis_quintile <- cut(PRS_data$PRS_osteoporosis, quantile(PRS_data$PRS_osteoporosis, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_osteoporosis_quintile)
PRS_data$PRS_osteoporosis_quintile <- factor(PRS_data$PRS_osteoporosis_quintile, ordered = FALSE)
PRS_data$PRS_osteoporosis_quintile <- relevel(PRS_data$PRS_osteoporosis_quintile, ref = "(-4.49,-0.831]")

# ovarian cancer quintiles
PRS_data$PRS_ovarian_quintile <- NA
PRS_data$PRS_ovarian_quintile <- cut(PRS_data$PRS_ovarian, quantile(PRS_data$PRS_ovarian, probs = seq(0, 1, 0.20), na.rm = TRUE))
table(PRS_data$PRS_ovarian_quintile)
PRS_data$PRS_ovarian_quintile <- factor(PRS_data$PRS_ovarian_quintile, ordered = FALSE)
PRS_data$PRS_ovarian_quintile <- relevel(PRS_data$PRS_ovarian_quintile, ref = "(-5.06,-1.07]")

```

## Merge all datasets

```{r merge_disease_data}

### all data
# merge in PCA data
for (k in seq_along(all_data)) {
    all_data[[k]] <- merge(
        all_data[[k]],
        pca_dfs[[k]],
        by = "eid",
        sort = FALSE)
}

# merge ICD-10 data with ACM datasets
all_data_ncd <- lapply(all_data, function(x)
    as.data.frame(
        merge(x, 
              dat_NCD_all,
              by = "eid",
              all.x = TRUE,
              all.y = FALSE,
              sort = FALSE)
    )
)

# merge baseline dx data with ACM datasets
all_data_ncd <- lapply(all_data_ncd, function(x)
    as.data.frame(
        merge(x, 
              dat_dx,
              by = "eid",
              all.x = TRUE,
              all.y = FALSE,
              sort = FALSE)
    )
)

# merge PRS data with ACM datasets
all_data_ncd <- lapply(all_data_ncd, function(x)
    as.data.frame(
        merge(x, 
              PRS_data,
              by = "eid",
              all.x = TRUE,
              all.y = FALSE,
              sort = FALSE)
    )
)

### scotwales

# merge in PCA data
for (k in seq_along(scotwales)) {
    scotwales[[k]] <- merge(
        scotwales[[k]],
        pca_dfs_scotwales[[k]],
        by = "eid",
        sort = FALSE)
}

# merge ICD-10 data with ACM datasets
scotwales_ncd <- lapply(scotwales, function(x)
    as.data.frame(
        merge(x, 
              dat_NCD_all,
              by = "eid",
              all.x = TRUE,
              all.y = FALSE,
              sort = FALSE)
    )
)

# merge baseline dx data with ACM datasets
scotwales_ncd <- lapply(scotwales_ncd, function(x)
    as.data.frame(
        merge(x, 
              dat_dx,
              by = "eid",
              all.x = TRUE,
              all.y = FALSE,
              sort = FALSE)
    )
)

# merge PRS data with ACM datasets
scotwales_ncd <- lapply(scotwales_ncd, function(x)
    as.data.frame(
        merge(x, 
              PRS_data,
              by = "eid",
              all.x = TRUE,
              all.y = FALSE,
              sort = FALSE)
    )
)



```

```{r remove_objects, eval=TRUE}
rm(
    all_data,
    scotwales,
    dat_dx,
    dat_NCD_all
)

invisible(gc())
```

## Create lists of disease names, indicator variables, survival time variables

```{r disease_list}

disease_list <- c(
    "All-cause mortality",
    "Colorectal cancer",
    "Lung cancer",
    "Esophageal cancer",
    "Liver cancer",
    "Pancreatic cancer",
    "Brain cancer",
    "Leukemia",
    "Lymphoma",
    "Breast cancer",
    "Ovarian cancer",
    "Prostate cancer",
    "Type II diabetes",
    "Ischemic heart disease",
    "Cerebrovascular diseases",
    "Emphysema, COPD",
    "Chronic liver diseases",
    "Chronic kidney diseases",
    "All-cause dementia",
    "Vascular dementia",
    "Alzheimer's disease",
    "Parkinson's disease",
    "Rheumatoid arthritis",
    "Macular degeneration",
    "Osteoporosis",
    "Osteoarthritis"
)

indicator <- c(
    Cs(
        ACM_event_indicator,
        colon_cancer_event,
        lung_cancer_event,
        eso_cancer_event,
        liver_cancer_event,
        pancreatic_cancer_event,
        brain_cancer_event,
        leukemia_event,
        lymphoma_event,
        breast_cancer_event,
        ovarian_cancer_event,
        prostate_cancer_event,
        diabetes_event,
        CVD_event,
        cerebro_event,
        COPD_event,
        liver_event,
        kidney_event,
        all_dementia_event,
        vasc_dementia_event,
        alzheimers_event,
        parkinsons_event,
        rheumatoid_event,
        macular_event,
        osteoporosis_event,
        osteoarthritis_event
    )
)

surv_time <- c(
    Cs(
        ACM_survival_time,
        colon_cancer_survival_time,
        lung_cancer_survival_time,
        eso_cancer_survival_time,
        liver_cancer_survival_time,
        pancreatic_cancer_survival_time,
        brain_cancer_survival_time,
        leukemia_survival_time,
        lymphoma_survival_time,
        breast_cancer_survival_time,
        ovarian_cancer_survival_time,
        prostate_cancer_survival_time,
        diabetes_survival_time,
        CVD_survival_time,
        cerebro_survival_time,
        COPD_survival_time,
        liver_survival_time,
        kidney_survival_time,
        all_dementia_survival_time,
        vasc_dementia_survival_time,
        alzheimers_survival_time,
        parkinsons_survival_time,
        rheumatoid_survival_time,
        macular_survival_time,
        osteoporosis_survival_time,
        osteoarthritis_survival_time
    )
)



```

## Create lists of survival objects for models

```{r surv_list}

# list of survival objects

surv_list <- c(
    "Surv(ACM_survival_time, ACM_event_indicator) ~
                  age_scaled + sex +   ",
    "Surv(colon_cancer_survival_time, colon_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(lung_cancer_survival_time, lung_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(eso_cancer_survival_time, eso_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(liver_cancer_survival_time, liver_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(pancreatic_cancer_survival_time, pancreatic_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(brain_cancer_survival_time, brain_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(leukemia_survival_time, leukemia_event) ~
                  age_scaled + sex +   ",
    "Surv(lymphoma_survival_time, lymphoma_event) ~
                  age_scaled + sex +   ",
    "Surv(breast_cancer_survival_time, breast_cancer_event) ~
                  age_scaled +   ",
    "Surv(ovarian_cancer_survival_time, ovarian_cancer_event) ~
                  age_scaled +   ",
    "Surv(prostate_cancer_survival_time, prostate_cancer_event) ~
                  age_scaled +   ",
    "Surv(diabetes_survival_time, diabetes_event) ~
                  age_scaled + sex +   ",
    "Surv(CVD_survival_time, CVD_event) ~
                  age_scaled + sex +   ",
    "Surv(cerebro_survival_time, cerebro_event) ~
                  age_scaled + sex +   ",
    "Surv(COPD_survival_time, COPD_event) ~
                  age_scaled + sex +   ",
    "Surv(liver_survival_time, liver_event) ~
                  age_scaled + sex +   ",
    "Surv(kidney_survival_time, kidney_event) ~
                  age_scaled + sex +   ",
    "Surv(all_dementia_survival_time, all_dementia_event) ~
                  age_scaled + sex +   ",
    "Surv(vasc_dementia_survival_time, vasc_dementia_event) ~
                  age_scaled + sex +   ",
    "Surv(alzheimers_survival_time, alzheimers_event) ~
                  age_scaled + sex +   ",
    "Surv(parkinsons_survival_time, parkinsons_event) ~
                  age_scaled + sex +   ",
    "Surv(rheumatoid_survival_time, rheumatoid_event) ~
                  age_scaled + sex +   ",
    "Surv(macular_survival_time, macular_event) ~
                  age_scaled + sex +   ",
    "Surv(osteoporosis_survival_time, osteoporosis_event) ~
                  age_scaled + sex +   ",
    "Surv(osteoarthritis_survival_time, osteoarthritis_event) ~
                  age_scaled + sex +   "
    
)

# duplicate
surv_list_copy <- surv_list

surv_list_tt <- c(
    "Surv(Start_time, Stop_time, ACM_event_indicator) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, colon_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, lung_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, eso_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, liver_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, pancreatic_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, brain_cancer_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, leukemia_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, lymphoma_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, breast_cancer_event) ~
                  age_scaled +   ",
    "Surv(Start_time, Stop_time, ovarian_cancer_event) ~
                  age_scaled +   ",
    "Surv(Start_time, Stop_time, prostate_cancer_event) ~
                  age_scaled +   ",
    "Surv(Start_time, Stop_time, diabetes_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, CVD_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, cerebro_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, COPD_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, liver_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, kidney_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, all_dementia_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, vasc_dementia_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, alzheimers_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, parkinsons_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, rheumatoid_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, macular_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, osteoporosis_event) ~
                  age_scaled + sex +   ",
    "Surv(Start_time, Stop_time, osteoarthritis_event) ~
                  age_scaled + sex +   "
)

surv_list_age_sex <- c(
    "Surv(ACM_survival_time, ACM_event_indicator) ~
                  age_scaled + sex",
    "Surv(colon_cancer_survival_time, colon_cancer_event) ~
                  age_scaled + sex",
    "Surv(lung_cancer_survival_time, lung_cancer_event) ~
                  age_scaled + sex",
    "Surv(eso_cancer_survival_time, eso_cancer_event) ~
                  age_scaled + sex",
    "Surv(liver_cancer_survival_time, liver_cancer_event) ~
                  age_scaled + sex",
    "Surv(pancreatic_cancer_survival_time, pancreatic_cancer_event) ~
                  age_scaled + sex",
    "Surv(brain_cancer_survival_time, brain_cancer_event) ~
                  age_scaled + sex",
    "Surv(leukemia_survival_time, leukemia_event) ~
                  age_scaled + sex",
    "Surv(lymphoma_survival_time, lymphoma_event) ~
                  age_scaled + sex",
    "Surv(breast_cancer_survival_time, breast_cancer_event) ~
                  age_scaled",
    "Surv(ovarian_cancer_survival_time, ovarian_cancer_event) ~
                  age_scaled",
    "Surv(prostate_cancer_survival_time, prostate_cancer_event) ~
                  age_scaled",
    "Surv(diabetes_survival_time, diabetes_event) ~
                  age_scaled + sex",
    "Surv(CVD_survival_time, CVD_event) ~
                  age_scaled + sex",
    "Surv(cerebro_survival_time, cerebro_event) ~
                  age_scaled + sex",
    "Surv(COPD_survival_time, COPD_event) ~
                  age_scaled + sex",
    "Surv(liver_survival_time, liver_event) ~
                  age_scaled + sex",
    "Surv(kidney_survival_time, kidney_event) ~
                  age_scaled + sex",
    "Surv(all_dementia_survival_time, all_dementia_event) ~
                  age_scaled + sex",
    "Surv(vasc_dementia_survival_time, vasc_dementia_event) ~
                  age_scaled + sex",
    "Surv(alzheimers_survival_time, alzheimers_event) ~
                  age_scaled + sex",
    "Surv(parkinsons_survival_time, parkinsons_event) ~
                  age_scaled + sex",
    "Surv(rheumatoid_survival_time, rheumatoid_event) ~
                  age_scaled + sex",
    "Surv(macular_survival_time, macular_event) ~
                  age_scaled + sex",
    "Surv(osteoporosis_survival_time, osteoporosis_event) ~
                  age_scaled + sex",
    "Surv(osteoarthritis_survival_time, osteoarthritis_event) ~
                  age_scaled + sex"
)


```

## Create list of significant exposures for each disease

```{r exposures_list}

exposures_list <- as.list(seq(1, length(disease_list)))

exposures_list[[1]] <- exposures

for (k in 2:length(exposures_list)) {
    sig_vars <- disease_results[[k-1]]
    exposures_list[[k]] <- unique(sig_vars$Variable[which(sig_vars$FDR < 0.05)])
    
}

names(exposures_list) <- disease_list

```

## Make list of functions to exclude prevalent cases for each disease

```{r prevalent_cases_list}

identify_prev_cases_list <- as.list(seq(1, length(disease_list)))

# mortality (no subset)
identify_prev_cases_list[[1]] <- function(x) {x <- x}

# colorectal cancer
identify_prev_cases_list[[2]] <- function(x) {
  
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$colon_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_colon_cancer <- NA
    x$baseline_colon_cancer[baseline_ncd] <- 1
    x$baseline_colon_cancer[which(x$colon_cancer_baseline_dx == 1)] <- 1
    x$baseline_colon_cancer[which(is.na(x$baseline_colon_cancer))] <- 0
    x$baseline_colon_cancer <- factor(x$baseline_colon_cancer)
    
    x <- x[which(x$baseline_colon_cancer == 0), ]
    return(x)
}

# lung cancer
identify_prev_cases_list[[3]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$lung_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_lung_cancer <- NA
    x$baseline_lung_cancer[baseline_ncd] <- 1
    x$baseline_lung_cancer[which(x$lung_cancer_baseline_dx == 1)] <- 1
    x$baseline_lung_cancer[which(is.na(x$baseline_lung_cancer))] <- 0
    x$baseline_lung_cancer <- factor(x$baseline_lung_cancer)
    
    x <- x[which(x$baseline_lung_cancer == 0), ]
    return(x)
}

# eso cancer
identify_prev_cases_list[[4]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$eso_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_eso_cancer <- NA
    x$baseline_eso_cancer[baseline_ncd] <- 1
    x$baseline_eso_cancer[which(x$eso_cancer_baseline_dx == 1)] <- 1
    x$baseline_eso_cancer[which(is.na(x$baseline_eso_cancer))] <- 0
    x$baseline_eso_cancer <- factor(x$baseline_eso_cancer)
    
    x <- x[which(x$baseline_eso_cancer == 0), ]
    return(x)
}

# liver cancer
identify_prev_cases_list[[5]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$liver_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_liver_cancer <- NA
    x$baseline_liver_cancer[baseline_ncd] <- 1
    x$baseline_liver_cancer[which(x$liver_cancer_baseline_dx == 1)] <- 1
    x$baseline_liver_cancer[which(is.na(x$baseline_liver_cancer))] <- 0
    x$baseline_liver_cancer <- factor(x$baseline_liver_cancer)
    
    x <- x[which(x$baseline_liver_cancer == 0), ]
    return(x)
}

# pancreatic cancer
identify_prev_cases_list[[6]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$pancreatic_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_pancreatic_cancer <- NA
    x$baseline_pancreatic_cancer[baseline_ncd] <- 1
    x$baseline_pancreatic_cancer[which(x$pancreatic_cancer_baseline_dx == 1)] <- 1
    x$baseline_pancreatic_cancer[which(is.na(x$baseline_pancreatic_cancer))] <- 0
    x$baseline_pancreatic_cancer <- factor(x$baseline_pancreatic_cancer)
    
    x <- x[which(x$baseline_pancreatic_cancer == 0), ]
    return(x)
}

# brain cancer
identify_prev_cases_list[[7]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$brain_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_brain_cancer <- NA
    x$baseline_brain_cancer[baseline_ncd] <- 1
    x$baseline_brain_cancer[which(x$brain_cancer_baseline_dx == 1)] <- 1
    x$baseline_brain_cancer[which(is.na(x$baseline_brain_cancer))] <- 0
    x$baseline_brain_cancer <- factor(x$baseline_brain_cancer)
    
    x <- x[which(x$baseline_brain_cancer == 0), ]
    return(x)
}

# leukemia
identify_prev_cases_list[[8]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$leukemia_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_leukemia <- NA
    x$baseline_leukemia[baseline_ncd] <- 1
    x$baseline_leukemia[which(x$leukemia_baseline_dx == 1)] <- 1
    x$baseline_leukemia[which(is.na(x$baseline_leukemia))] <- 0
    x$baseline_leukemia <- factor(x$baseline_leukemia)
    
    x <- x[which(x$baseline_leukemia == 0), ]
    return(x)
}

# lymphoma
identify_prev_cases_list[[9]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$lymphoma_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_lymphoma <- NA
    x$baseline_lymphoma[baseline_ncd] <- 1
    x$baseline_lymphoma[which(x$lymphoma_baseline_dx == 1)] <- 1
    x$baseline_lymphoma[which(is.na(x$baseline_lymphoma))] <- 0
    x$baseline_lymphoma <- factor(x$baseline_lymphoma)
    
    x <- x[which(x$baseline_lymphoma == 0), ]
    return(x)
}

# breast cancer
identify_prev_cases_list[[10]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$breast_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_breast_cancer <- NA
    x$baseline_breast_cancer[baseline_ncd] <- 1
    x$baseline_breast_cancer[which(x$breast_cancer_baseline_dx == 1)] <- 1
    x$baseline_breast_cancer[which(is.na(x$baseline_breast_cancer))] <- 0
    x$baseline_breast_cancer <- factor(x$baseline_breast_cancer)
    
    x <- x[which(x$baseline_breast_cancer == 0), ]
    x <- x[which(x$sex == "Female"), ]
    return(x)
}

# ovarian cancer
identify_prev_cases_list[[11]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$ovarian_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_ovarian_cancer <- NA
    x$baseline_ovarian_cancer[baseline_ncd] <- 1
    x$baseline_ovarian_cancer[which(x$ovarian_cancer_baseline_dx == 1)] <- 1
    x$baseline_ovarian_cancer[which(is.na(x$baseline_ovarian_cancer))] <- 0
    x$baseline_ovarian_cancer <- factor(x$baseline_ovarian_cancer)
    
    x <- x[which(x$baseline_ovarian_cancer == 0), ]
    x <- x[which(x$sex == "Female"), ]
    return(x)
}

# prostate cancer
identify_prev_cases_list[[12]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$prostate_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_prostate_cancer <- NA
    x$baseline_prostate_cancer[baseline_ncd] <- 1
    x$baseline_prostate_cancer[which(x$prostate_cancer_baseline_dx == 1)] <- 1
    x$baseline_prostate_cancer[which(is.na(x$baseline_prostate_cancer))] <- 0
    x$baseline_prostate_cancer <- factor(x$baseline_prostate_cancer)
    
    x <- x[which(x$baseline_prostate_cancer == 0), ]
    x <- x[which(x$sex == "Male"), ]
    return(x)
}

# diabetes
identify_prev_cases_list[[13]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$diabetes_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_diabetes <- NA
    x$baseline_diabetes[baseline_ncd] <- 1
    x$baseline_diabetes[which(x$diabetes_diagnosis == "Yes")] <- 1
    x$baseline_diabetes[which(x$diabetes_baseline_dx == 1)] <- 1
    x$baseline_diabetes[which(x$insulin == "Yes")] <- 1
    x$baseline_diabetes[which(x$hbA1c >= 48)] <- 1
    x$baseline_diabetes[which(x$glucose >= 11.1)] <- 1
    x$baseline_diabetes[which(is.na(x$baseline_diabetes))] <- 0
    x$baseline_diabetes <- factor(x$baseline_diabetes)
    
    x <- x[which(x$baseline_diabetes == 0), ]
    return(x)
}

# cvd
identify_prev_cases_list[[14]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$cvd_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_cvd <- NA
    x$baseline_cvd[baseline_ncd] <- 1
    x$baseline_cvd[which(x$heart_attack_diagnosis == "Yes")] <- 1
    x$baseline_cvd[which(x$angina_diagnosis == "Yes")] <- 1
    x$baseline_cvd[which(x$CVD_baseline_dx == 1)] <- 1
    x$baseline_cvd[which(is.na(x$baseline_cvd))] <- 0
    x$baseline_cvd <- factor(x$baseline_cvd)
    
    x <- x[which(x$baseline_cvd == 0), ]
    return(x)
}

# cerebro
identify_prev_cases_list[[15]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$cerebro_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_cerebro <- NA
    x$baseline_cerebro[baseline_ncd] <- 1
    x$baseline_cerebro[which(x$stroke_diagnosis == "Yes")] <- 1
    x$baseline_cerebro[which(x$cerebro_baseline_dx == 1)] <- 1
    x$baseline_cerebro[which(is.na(x$baseline_cerebro))] <- 0
    x$baseline_cerebro <- factor(x$baseline_cerebro)
    
    x <- x[which(x$baseline_cerebro == 0), ]
    return(x)
}

# copd
identify_prev_cases_list[[16]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$copd_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_copd <- NA
    x$baseline_copd[baseline_ncd] <- 1
    x$baseline_copd[which(x$bronchitis_emphysema_diagnosis == "Yes")] <- 1
    x$baseline_copd[which(x$COPD_baseline_dx == 1)] <- 1
    x$baseline_copd[which(is.na(x$baseline_copd))] <- 0
    x$baseline_copd <- factor(x$baseline_copd)
    
    x <- x[which(x$baseline_copd == 0), ]
    return(x)
}

# liver
identify_prev_cases_list[[17]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$liver_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_liver <- NA
    x$baseline_liver[baseline_ncd] <- 1
    x$baseline_liver[which(x$liver_baseline_dx == 1)] <- 1
    x$baseline_liver[which(is.na(x$baseline_liver))] <- 0
    x$baseline_liver <- factor(x$baseline_liver)
    
    x <- x[which(x$baseline_liver == 0), ]
    return(x)
}

# kidney
identify_prev_cases_list[[18]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$kidney_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_kidney <- NA
    x$baseline_kidney[baseline_ncd] <- 1
    x$baseline_kidney[which(x$kidney_baseline_dx == 1)] <- 1
    x$baseline_kidney[which(is.na(x$baseline_kidney))] <- 0
    x$baseline_kidney <- factor(x$baseline_kidney)
    
    x <- x[which(x$baseline_kidney == 0), ]
    return(x)
}

# all_dementia
identify_prev_cases_list[[19]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$all_dementia_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_all_dementia <- NA
    x$baseline_all_dementia[baseline_ncd] <- 1
    x$baseline_all_dementia[which(x$all_dementia_baseline_dx == 1)] <- 1
    x$baseline_all_dementia[which(is.na(x$baseline_all_dementia))] <- 0
    x$baseline_all_dementia <- factor(x$baseline_all_dementia)
    
    x <- x[which(x$baseline_all_dementia == 0), ]
    return(x)
}

# vasc_dementia
identify_prev_cases_list[[20]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$vasc_dementia_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_vasc_dementia <- NA
    x$baseline_vasc_dementia[baseline_ncd] <- 1
    x$baseline_vasc_dementia[which(x$vasc_dementia_baseline_dx == 1)] <- 1
    x$baseline_vasc_dementia[which(is.na(x$baseline_vasc_dementia))] <- 0
    x$baseline_vasc_dementia <- factor(x$baseline_vasc_dementia)
    
    x <- x[which(x$baseline_vasc_dementia == 0), ]
    return(x)
}

# alzheimers
identify_prev_cases_list[[21]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$alzheimers_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_alzheimers <- NA
    x$baseline_alzheimers[baseline_ncd] <- 1
    x$baseline_alzheimers[which(x$alzheimers_baseline_dx == 1)] <- 1
    x$baseline_alzheimers[which(is.na(x$baseline_alzheimers))] <- 0
    x$baseline_alzheimers <- factor(x$baseline_alzheimers)
    
    x <- x[which(x$baseline_alzheimers == 0), ]
    return(x)
}

# parkinsons
identify_prev_cases_list[[22]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$parkinsons_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_parkinsons <- NA
    x$baseline_parkinsons[baseline_ncd] <- 1
    x$baseline_parkinsons[which(x$parkinsons_baseline_dx == 1)] <- 1
    x$baseline_parkinsons[which(is.na(x$baseline_parkinsons))] <- 0
    x$baseline_parkinsons <- factor(x$baseline_parkinsons)
    
    x <- x[which(x$baseline_parkinsons == 0), ]
    return(x)
}

# rheumatoid
identify_prev_cases_list[[23]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$rheumatoid_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_rheumatoid <- NA
    x$baseline_rheumatoid[baseline_ncd] <- 1
    x$baseline_rheumatoid[which(x$rheumatoid_baseline_dx == 1)] <- 1
    x$baseline_rheumatoid[which(is.na(x$baseline_rheumatoid))] <- 0
    x$baseline_rheumatoid <- factor(x$baseline_rheumatoid)
    
    x <- x[which(x$baseline_rheumatoid == 0), ]
    return(x)
}

# macular
identify_prev_cases_list[[24]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$macular_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_macular <- NA
    x$baseline_macular[baseline_ncd] <- 1
    x$baseline_macular[which(x$macular_baseline_dx == 1)] <- 1
    x$baseline_macular[which(is.na(x$baseline_macular))] <- 0
    x$baseline_macular <- factor(x$baseline_macular)
    
    x <- x[which(x$baseline_macular == 0), ]
    return(x)
}

# osteoporosis
identify_prev_cases_list[[25]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$osteoporosis_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_osteoporosis <- NA
    x$baseline_osteoporosis[baseline_ncd] <- 1
    x$baseline_osteoporosis[which(x$osteoporosis_baseline_dx == 1)] <- 1
    x$baseline_osteoporosis[which(is.na(x$baseline_osteoporosis))] <- 0
    x$baseline_osteoporosis <- factor(x$baseline_osteoporosis)
    
    x <- x[which(x$baseline_osteoporosis == 0), ]
    return(x)
}

# osteoarthritis
identify_prev_cases_list[[26]] <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$osteoarthritis_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_osteoarthritis <- NA
    x$baseline_osteoarthritis[baseline_ncd] <- 1
    x$baseline_osteoarthritis[which(x$osteoarthritis_baseline_dx == 1)] <- 1
    x$baseline_osteoarthritis[which(is.na(x$baseline_osteoarthritis))] <- 0
    x$baseline_osteoarthritis <- factor(x$baseline_osteoarthritis)
    
    x <- x[which(x$baseline_osteoarthritis == 0), ]
    return(x)
}



```

## Make lists of PRS diseases and exposures for PRS models

```{r PRS_prep}

# diseases where we have PRS data
PRS_disease_list <- c(
    "All-cause mortality",
    "Colorectal cancer",
    "Lung cancer",
    "Esophageal cancer",
    "Pancreatic cancer",
    "Leukemia",
    "Breast cancer",
    "Ovarian cancer",
    "Prostate cancer",
    "Type II diabetes",
    "Ischemic heart disease",
    "Cerebrovascular diseases",
    "Emphysema, COPD",
    "All-cause dementia",
    "Vascular dementia",
    "Alzheimer's disease",
    "Parkinson's disease",
    "Rheumatoid arthritis",
    "Macular degeneration",
    "Osteoporosis"
)


# list of PRS vars
PRS_list <- c(
    "GeP_PC1",
    "GeP_PC2",
    "GeP_PC3",
    "GeP_PC4",
    "GeP_Array",
    "PRS_breast_quintile",
    "PRS_bowel_quintile",
    "PRS_prostate_quintile",
    "PRS_T2D_quintile",
    "PRS_CVD_quintile",
    "PRS_CAD_quintile",
    "PRS_stroke_quintile",
    "PRS_COPD_quintile",
    "PRS_alzheimers_quintile",
    "PRS_parkinsons_quintile",
    "PRS_rheumatoid_quintile",
    "PRS_macular_quintile",
    "PRS_osteoporosis_quintile",
    "PRS_ovarian_quintile",
    "PRS_leukemia_quintile",
    "PRS_lung_quintile",
    "PRS_eso_quintile",
    "PRS_pancreatic_quintile"
)

# list of PRS covariates
PRS_covars <- c(
    "GeP_PC1",
    "GeP_PC2",
    "GeP_PC3",
    "GeP_PC4",
    "GeP_Array"
)


# create exposure lists for each PRS model by adding PRS variable(s) to exposures
names(exposures_list) <- disease_list
PRS_exposures <- exposures_list
PRS_exposures <- PRS_exposures[which(names(PRS_exposures) %in% PRS_disease_list)]


PRS_exposures$`All-cause mortality` <- 
    c(exposures_list$`All-cause mortality`, PRS_list)

PRS_exposures$`Colorectal cancer` <- 
    c(exposures_list$`Colorectal cancer`, c(PRS_covars, "PRS_bowel_quintile"))

PRS_exposures$`Breast cancer` <- 
    c(exposures_list$`Breast cancer`, c(PRS_covars, "PRS_breast_quintile"))

PRS_exposures$`Ovarian cancer` <-
    c(exposures_list$`Ovarian cancer`, c(PRS_covars, "PRS_ovarian_quintile"))

PRS_exposures$`Prostate cancer` <- 
    c(exposures_list$`Prostate cancer`, c(PRS_covars, "PRS_prostate_quintile"))

PRS_exposures$`Type II diabetes` <- 
    c(exposures_list$`Type II diabetes`, c(PRS_covars, "PRS_T2D_quintile"))

PRS_exposures$`Ischemic heart disease` <- 
    c(exposures_list$`Ischemic heart disease`, c(PRS_covars, "PRS_CVD_quintile", "PRS_CAD_quintile"))

PRS_exposures$`Cerebrovascular diseases` <- 
    c(exposures_list$`Cerebrovascular diseases`, c(PRS_covars, "PRS_stroke_quintile"))

PRS_exposures$`Emphysema, COPD` <- 
    c(exposures_list$`Emphysema, COPD`, c(PRS_covars, "PRS_COPD_quintile"))

PRS_exposures$`All-cause dementia` <- 
    c(exposures_list$`All-cause dementia`, c(PRS_covars, "PRS_alzheimers_quintile"))

PRS_exposures$`Vascular dementia` <- 
    c(exposures_list$`Vascular dementia`, c(PRS_covars, "PRS_alzheimers_quintile"))

PRS_exposures$`Alzheimer's disease` <- 
    c(exposures_list$`Alzheimer's disease`, c(PRS_covars, "PRS_alzheimers_quintile"))

PRS_exposures$`Parkinson's disease` <- 
    c(exposures_list$`Parkinson's disease`, c(PRS_covars, "PRS_parkinsons_quintile"))

PRS_exposures$`Rheumatoid arthritis` <- 
    c(exposures_list$`Rheumatoid arthritis`, c(PRS_covars, "PRS_rheumatoid_quintile"))

PRS_exposures$`Macular degeneration` <- 
    c(exposures_list$`Macular degeneration`, c(PRS_covars, "PRS_macular_quintile"))

PRS_exposures$Osteoporosis <- 
    c(exposures_list$Osteoporosis, c(PRS_covars, "PRS_osteoporosis_quintile"))

PRS_exposures$`Lung cancer` <- 
    c(exposures_list$`Lung cancer`, c(PRS_covars, "PRS_lung_quintile"))

PRS_exposures$Leukemia <- 
    c(exposures_list$Leukemia, c(PRS_covars, "PRS_leukemia_quintile"))

PRS_exposures$`Esophageal cancer` <- 
    c(exposures_list$`Esophageal cancer`, c(PRS_covars, "PRS_eso_quintile"))

PRS_exposures$`Pancreatic cancer` <- 
    c(exposures_list$`Pancreatic cancer`, c(PRS_covars, "PRS_pancreatic_quintile"))

```

## Run crosstab checks between exposures and each outcome

# Exposome exposures

```{r crosstab_check}

library(caret)

sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)

problem_vars <- as.list(rep(1, length(exposures_list)))

for (k in sig_exposures) {
    
    
    # subset the data
    test_data <- all_data_ncd[[1]]
    test_data <- identify_prev_cases_list[[k]](test_data)
    test_data <- test_data[c(exposures_list[[k]], indicator[k])]
    test_data <- na.omit(test_data)
    
    ### crosstab check
    
    # convert mortality event indicator to factor for crosstabs
    test_data[[indicator[k]]] <- factor(test_data[[indicator[k]]])
    
    # identify all factor variables
    vars <- names(test_data[exposures_list[[k]]])
    fact_vars <- sapply(test_data[exposures_list[[k]]], is.factor)
    trial_vars <- vars[fact_vars]
    
    # subset dataset to just factors
    crosstab_data <- test_data[c(trial_vars, indicator[k])]
    
    # get column number for event indicator
    mort_index <- which(colnames(crosstab_data) == indicator[k])
    
    # create list for crosstab of results
    tables <- as.list(rep(1, ncol(crosstab_data)-1))
    
    # create list of column positions
    col_position <- seq(1, ncol(crosstab_data))
    
    # remove mortality event column from list of columns
    col_position <- col_position[which(col_position != mort_index)]
    
    # run loop of crosstabs between variables and mortality
    for (j in col_position) {
        tables[[j]] <- 
            as.data.frame.matrix(table(crosstab_data[[j]], crosstab_data[[indicator[k]]]))
        tables[[j]][3] <- names(crosstab_data[j])
    }
    
    # rbind all results together into single table
    crosstabs <- Reduce(function(x, y) 
        rbind(x = x, 
              y = y), 
        tables)
    
    # make column for variable levels
    crosstabs$levels <- NA
    
    # add in levels to each variable
    for (j in col_position) {
        crosstabs$levels[which(crosstabs$V3 == names(crosstab_data[j]))] <- levels(crosstab_data[[j]])
    }
    
    # fix rownames
    rownames(crosstabs) <- NULL
    crosstabs <- crosstabs[which(crosstabs$V3 != 1), ]
    crosstabs <- crosstabs[which(crosstabs$V3 != "sex"), ]
    
    
    # get list of variables with <30 mortality cases in a response level
    low_variance <- crosstabs$V3[which(crosstabs$'1' < 10)]
    
    # get a df of these problem vars that shows all response categories and the cross-tab tables
    problem_vars[[k]] <- crosstabs[which(crosstabs$V3 %in% low_variance), ]
    
    cat("\r", paste("completed:", paste(k, length(problem_vars), sep = "/")))
}

```

# Exposome + PRS exposures

```{r crosstab_check_PRS}

library(caret)

sig_exposures_prs <- sapply(PRS_exposures, length)
sig_exposures_prs <- which(sig_exposures_prs != 0)

problem_vars_prs <- as.list(rep(1, length(PRS_exposures)))

for (k in sig_exposures_prs) {
    
    
    # subset the data
    test_data <- all_data_ncd[[1]]
    test_data <- identify_prev_cases_list[[k]](test_data)
    test_data <- test_data[c(PRS_exposures[[k]], indicator[k])]
    test_data <- na.omit(test_data)
    
    ### crosstab check
    
    # convert mortality event indicator to factor for crosstabs
    test_data[[indicator[k]]] <- factor(test_data[[indicator[k]]])
    
    # identify all factor variables
    vars <- names(test_data[PRS_exposures[[k]]])
    fact_vars <- sapply(test_data[PRS_exposures[[k]]], is.factor)
    trial_vars <- vars[fact_vars]
    
    # subset dataset to just factors
    crosstab_data <- test_data[c(trial_vars, indicator[k])]
    
    # get column number for event indicator
    mort_index <- which(colnames(crosstab_data) == indicator[k])
    
    # create list for crosstab of results
    tables <- as.list(rep(1, ncol(crosstab_data)-1))
    
    # create list of column positions
    col_position <- seq(1, ncol(crosstab_data))
    
    # remove mortality event column from list of columns
    col_position <- col_position[which(col_position != mort_index)]
    
    # run loop of crosstabs between variables and mortality
    for (j in col_position) {
        tables[[j]] <- 
            as.data.frame.matrix(table(crosstab_data[[j]], crosstab_data[[indicator[k]]]))
        tables[[j]][3] <- names(crosstab_data[j])
    }
    
    # rbind all results together into single table
    crosstabs <- Reduce(function(x, y) 
        rbind(x = x, 
              y = y), 
        tables)
    
    # make column for variable levels
    crosstabs$levels <- NA
    
    # add in levels to each variable
    for (j in col_position) {
        crosstabs$levels[which(crosstabs$V3 == names(crosstab_data[j]))] <- levels(crosstab_data[[j]])
    }
    
    # fix rownames
    rownames(crosstabs) <- NULL
    crosstabs <- crosstabs[which(crosstabs$V3 != 1), ]
    crosstabs <- crosstabs[which(crosstabs$V3 != "sex"), ]
    
    
    # get list of variables with <30 mortality cases in a response level
    low_variance <- crosstabs$V3[which(crosstabs$'1' < 10)]
    
    # get a df of these problem vars that shows all response categories and the cross-tab tables
    problem_vars_prs[[k]] <- crosstabs[which(crosstabs$V3 %in% low_variance), ]
    
    cat("\r", paste("completed:", paste(k, length(problem_vars_prs), sep = "/")))
}

# subset to only models that were run
names(problem_vars_prs) <- names(PRS_exposures)

```

## Remove problem variables where there are too few cases per response level

```{r remove_problem_vars}

sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)

other <- which(names(exposures_list) %nin% names(PRS_exposures))
together <- which(names(exposures_list) %in% names(PRS_exposures))
together_vars <- exposures_list

exp_vars <- as.list(seq_along(problem_vars))
for (k in sig_exposures) {
    exp_vars[[k]] <- problem_vars[[k]]$V3
    exp_vars[[k]] <- exp_vars[[k]][exp_vars[[k]] != "accommodation_type"]
}

prs_vars <- as.list(seq_along(problem_vars_prs))
for (k in seq_along(prs_vars)) {
    prs_vars[[k]] <- problem_vars_prs[[k]]$V3
    prs_vars[[k]] <- prs_vars[[k]][prs_vars[[k]] != "accommodation_type"]
}

# remove problem variables for exposome only
for (k in other) {
    exposures_list[[k]] <- exposures_list[[k]][which(exposures_list[[k]] %nin% exp_vars[[k]])]
}

for (k in together) {
    exposures_list[[k]] <- exposures_list[[k]][which(exposures_list[[k]] %nin% prs_vars[[which(names(problem_vars_prs) == names(exposures_list)[k])]])]
}

# remove problem variables for PRS
for (k in seq_along(PRS_exposures)) {
    PRS_exposures[[k]] <- PRS_exposures[[k]][which(PRS_exposures[[k]] %nin% prs_vars[[k]])]
}

```

## Recode variables according to crosstabs

```{r crosstab_recode}

crosstab_recode <- function(x) {
  # accommodation_type
  x$accommodation_type[which(x$accommodation_type == "Mobile or temporary structure (i.e. caravan)")] <- NA
  old_levels <- levels(x$accommodation_type)
  x$accommodation_type <- 
    factor(x$accommodation_type,
           ordered = FALSE,
           levels = old_levels[which(old_levels !="Mobile or temporary structure (i.e. caravan)")])
  
  return(x)
  
}

all_data_ncd <- lapply(all_data_ncd, crosstab_recode)
scotwales_ncd <- lapply(scotwales_ncd, crosstab_recode)

```

## Create exposure lists for clinical risk factor models

```{r clinical_vars}

# copy exposures list
clinical_vars <- exposures_list
# get index of outcomes for which we have PRS
together <- which(names(exposures_list) %in% names(PRS_exposures))
# copy list of PRS exposures to those outcomes
clinical_vars[together] <- PRS_exposures

## functions to code prevalent disease diagnoses to use as clinical risk factors

baseline_cvd <- function(x) {
    baseline_ncd <- which(x$cvd_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_cvd <- NA
    x$baseline_cvd[baseline_ncd] <- 1
    x$baseline_cvd[which(x$heart_attack_diagnosis == "Yes")] <- 1
    x$baseline_cvd[which(x$angina_diagnosis == "Yes")] <- 1
    x$baseline_cvd[which(x$CVD_baseline_dx == 1)] <- 1
    x$baseline_cvd[which(is.na(x$baseline_cvd))] <- 0
    x$baseline_cvd <- factor(x$baseline_cvd)
    
    return(x)
}

baseline_t2d <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$diabetes_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_diabetes <- NA
    x$baseline_diabetes[baseline_ncd] <- 1
    x$baseline_diabetes[which(x$diabetes_diagnosis == "Yes")] <- 1
    x$baseline_diabetes[which(x$diabetes_baseline_dx == 1)] <- 1
    x$baseline_diabetes[which(x$insulin == "Yes")] <- 1
    x$baseline_diabetes[which(x$hbA1c >= 48)] <- 1
    x$baseline_diabetes[which(x$glucose >= 11.1)] <- 1
    x$baseline_diabetes[which(is.na(x$baseline_diabetes))] <- 0
    x$baseline_diabetes <- factor(x$baseline_diabetes)
    
    return(x)
}

baseline_breast <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$breast_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_breast_cancer <- NA
    x$baseline_breast_cancer[baseline_ncd] <- 1
    x$baseline_breast_cancer[which(x$breast_cancer_baseline_dx == 1)] <- 1
    x$baseline_breast_cancer[which(is.na(x$baseline_breast_cancer))] <- 0
    x$baseline_breast_cancer <- factor(x$baseline_breast_cancer)
    
    return(x)
}

baseline_crc <- function(x) {
  
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$colon_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_colon_cancer <- NA
    x$baseline_colon_cancer[baseline_ncd] <- 1
    x$baseline_colon_cancer[which(x$colon_cancer_baseline_dx == 1)] <- 1
    x$baseline_colon_cancer[which(is.na(x$baseline_colon_cancer))] <- 0
    x$baseline_colon_cancer <- factor(x$baseline_colon_cancer)
    
    return(x)
}

baseline_leukemia <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$leukemia_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_leukemia <- NA
    x$baseline_leukemia[baseline_ncd] <- 1
    x$baseline_leukemia[which(x$leukemia_baseline_dx == 1)] <- 1
    x$baseline_leukemia[which(is.na(x$baseline_leukemia))] <- 0
    x$baseline_leukemia <- factor(x$baseline_leukemia)
    
    return(x)
}

baseline_lung <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$lung_cancer_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_lung_cancer <- NA
    x$baseline_lung_cancer[baseline_ncd] <- 1
    x$baseline_lung_cancer[which(x$lung_cancer_baseline_dx == 1)] <- 1
    x$baseline_lung_cancer[which(is.na(x$baseline_lung_cancer))] <- 0
    x$baseline_lung_cancer <- factor(x$baseline_lung_cancer)
    
    return(x)
}

baseline_copd <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$copd_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_copd <- NA
    x$baseline_copd[baseline_ncd] <- 1
    x$baseline_copd[which(x$bronchitis_emphysema_diagnosis == "Yes")] <- 1
    x$baseline_copd[which(x$COPD_baseline_dx == 1)] <- 1
    x$baseline_copd[which(is.na(x$baseline_copd))] <- 0
    x$baseline_copd <- factor(x$baseline_copd)
    
    return(x)
}

baseline_dyslipidemia <- function(x) {
  
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$dyslipidemia_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_dyslipidemia <- NA
    x$baseline_dyslipidemia[baseline_ncd] <- 1
    x$baseline_dyslipidemia[which(x$cholesterol_meds == "Yes")] <- 1
    x$baseline_dyslipidemia[which(x$cholesterol >= (240 / 38.67))] <- 1
    x$baseline_dyslipidemia[which(x$LDL_direct >= (160 / 38.67))] <- 1
    x$baseline_dyslipidemia[which(x$HDL_cholesterol < (40 / 38.67))] <- 1
    x$baseline_dyslipidemia[which(x$triglycerides >= (200 / 88.57))] <- 1
    x$baseline_dyslipidemia[which(is.na(x$baseline_dyslipidemia))] <- 0
    x$baseline_dyslipidemia <- factor(x$baseline_dyslipidemia)
    x$baseline_dyslipidemia <- relevel(x$baseline_dyslipidemia, ref = "0")
    
    return(x)
}

baseline_liver <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$liver_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_liver <- NA
    x$baseline_liver[baseline_ncd] <- 1
    x$baseline_liver[which(x$liver_baseline_dx == 1)] <- 1
    x$baseline_liver[which(is.na(x$baseline_liver))] <- 0
    x$baseline_liver <- factor(x$baseline_liver)
    
    return(x)
}

baseline_cerebro <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$cerebro_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_cerebro <- NA
    x$baseline_cerebro[baseline_ncd] <- 1
    x$baseline_cerebro[which(x$stroke_diagnosis == "Yes")] <- 1
    x$baseline_cerebro[which(x$cerebro_baseline_dx == 1)] <- 1
    x$baseline_cerebro[which(is.na(x$baseline_cerebro))] <- 0
    x$baseline_cerebro <- factor(x$baseline_cerebro)
    
    return(x)
}

baseline_rheumatoid <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$rheumatoid_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_rheumatoid <- NA
    x$baseline_rheumatoid[baseline_ncd] <- 1
    x$baseline_rheumatoid[which(x$rheumatoid_baseline_dx == 1)] <- 1
    x$baseline_rheumatoid[which(is.na(x$baseline_rheumatoid))] <- 0
    x$baseline_rheumatoid <- factor(x$baseline_rheumatoid)
    
    return(x)
}

baseline_kidney <- function(x) {
    
    # define all participants with NCD diagnosis before or equal to recruitment date
    baseline_ncd <- which(x$kidney_censor_date <= x$recruitment_date)
    
    # creating health indicator for baseline disease in total population
    x$baseline_kidney <- NA
    x$baseline_kidney[baseline_ncd] <- 1
    x$baseline_kidney[which(x$kidney_baseline_dx == 1)] <- 1
    x$baseline_kidney[which(is.na(x$baseline_kidney))] <- 0
    x$baseline_kidney <- factor(x$baseline_kidney)
    
    return(x)
}

## apply functions in each dataset
all_data_ncd <- lapply(all_data_ncd, baseline_cvd)
all_data_ncd <- lapply(all_data_ncd, baseline_t2d)
all_data_ncd <- lapply(all_data_ncd, baseline_breast)
all_data_ncd <- lapply(all_data_ncd, baseline_crc)
all_data_ncd <- lapply(all_data_ncd, baseline_leukemia)
all_data_ncd <- lapply(all_data_ncd, baseline_lung)
all_data_ncd <- lapply(all_data_ncd, baseline_copd)
all_data_ncd <- lapply(all_data_ncd, baseline_dyslipidemia)
all_data_ncd <- lapply(all_data_ncd, baseline_liver)
all_data_ncd <- lapply(all_data_ncd, baseline_cerebro)
all_data_ncd <- lapply(all_data_ncd, baseline_rheumatoid)
all_data_ncd <- lapply(all_data_ncd, baseline_kidney)

scotwales_ncd <- lapply(scotwales_ncd, baseline_cvd)
scotwales_ncd <- lapply(scotwales_ncd, baseline_t2d)
scotwales_ncd <- lapply(scotwales_ncd, baseline_breast)
scotwales_ncd <- lapply(scotwales_ncd, baseline_crc)
scotwales_ncd <- lapply(scotwales_ncd, baseline_leukemia)
scotwales_ncd <- lapply(scotwales_ncd, baseline_lung)
scotwales_ncd <- lapply(scotwales_ncd, baseline_copd)
scotwales_ncd <- lapply(scotwales_ncd, baseline_dyslipidemia)
scotwales_ncd <- lapply(scotwales_ncd, baseline_liver)
scotwales_ncd <- lapply(scotwales_ncd, baseline_cerebro)
scotwales_ncd <- lapply(scotwales_ncd, baseline_rheumatoid)
scotwales_ncd <- lapply(scotwales_ncd, baseline_kidney)

## add extra clinical risk factors to exposure list for each disease
# information on cancer risk factors was taken from cancer reasearch UK (https://www.cancerresearchuk.org/about-cancer/breast-cancer/risks-causes/risk-factors)

clinical_vars$`Cerebrovascular diseases` <- 
    c(clinical_vars$`Cerebrovascular diseases`, Cs(baseline_cvd, baseline_diabetes, blood_pressure_meds, systolic_bp, BMI, baseline_dyslipidemia))

clinical_vars$`All-cause dementia` <- 
    c(clinical_vars$`All-cause dementia`, Cs(baseline_diabetes, blood_pressure_meds, systolic_bp, BMI, hearing_difficulty, MDD, prospective_memory, identify_matches_mean_time, fluid_intelligence
))

clinical_vars$`Vascular dementia` <- 
    c(clinical_vars$`Vascular dementia`, Cs(baseline_diabetes, blood_pressure_meds, systolic_bp, BMI, hearing_difficulty, MDD, prospective_memory, identify_matches_mean_time, fluid_intelligence))

clinical_vars$`Alzheimer's disease` <- 
    c(clinical_vars$`Alzheimer's disease`, Cs(baseline_diabetes, blood_pressure_meds, systolic_bp, BMI, hearing_difficulty, MDD, prospective_memory, identify_matches_mean_time, fluid_intelligence))

clinical_vars$`Lung cancer` <- 
    c(clinical_vars$`Lung cancer`, Cs(FEV1_standardized, BMI, baseline_copd))

clinical_vars$`Emphysema, COPD` <- 
    c(clinical_vars$`Emphysema, COPD`, Cs(FEV1_standardized, BMI))

clinical_vars$`Ovarian cancer` <-
    c(clinical_vars$`Ovarian cancer`, Cs(BMI, HRT, hysterectomy, cancer_diagnosis))

clinical_vars$`Breast cancer` <- 
    c(clinical_vars$`Breast cancer`, Cs(IGF1, BMI, oral_contraceptive, HRT, menarche_age, number_births, stillbirth, cancer_diagnosis, standing_height, testosterone))

clinical_vars$`Prostate cancer` <- 
    c(clinical_vars$`Prostate cancer`, Cs(IGF1, BMI, standing_height))

clinical_vars$`Colorectal cancer` <- 
    c(clinical_vars$`Colorectal cancer`, Cs(BMI, baseline_diabetes))

clinical_vars$Leukemia <- 
    c(clinical_vars$Leukemia, Cs(BMI, baseline_diabetes, leukocyte_count))

clinical_vars$`Esophageal cancer` <- 
    c(clinical_vars$`Esophageal cancer`, Cs(BMI, omeprazole, ranitidine, hot_drink_temp)) 

clinical_vars$`Pancreatic cancer` <- 
    c(clinical_vars$`Pancreatic cancer`, Cs(BMI, baseline_diabetes, blood_pressure_meds, baseline_cvd, systolic_bp, baseline_dyslipidemia))

clinical_vars$`Liver cancer` <- 
    c(clinical_vars$`Liver cancer`, Cs(BMI, baseline_diabetes, baseline_liver, leukocyte_count))

clinical_vars$`Parkinson's disease` <- 
    c(clinical_vars$`Parkinson's disease`, Cs(baseline_cerebro))

clinical_vars$Osteoporosis <- 
    c(clinical_vars$Osteoporosis, Cs(heel_bone_mineral_density, testosterone, BMI, fracture_last_5yrs, baseline_rheumatoid, omeprazole, ranitidine, baseline_kidney, baseline_liver, cancer_diagnosis))

clinical_vars$`Rheumatoid arthritis` <- 
    c(clinical_vars$`Rheumatoid arthritis`, Cs(maternal_smoking, BMI))

clinical_vars$`Macular degeneration` <- 
    c(clinical_vars$`Macular degeneration`, Cs(BMI, baseline_cvd, blood_pressure_meds, systolic_bp))

clinical_vars$`Ischemic heart disease` <- 
    c(clinical_vars$`Ischemic heart disease`, Cs(blood_pressure_meds, systolic_bp, baseline_dyslipidemia, BMI, baseline_diabetes, MDD))

clinical_vars$`Type II diabetes` <- 
    c(clinical_vars$`Type II diabetes`, Cs(blood_pressure_meds, systolic_bp, baseline_dyslipidemia, BMI, MDD))

clinical_vars$`Chronic liver diseases` <- 
    c(clinical_vars$`Chronic liver diseases`, Cs(BMI, baseline_diabetes, cancer_diagnosis, leukocyte_count))

clinical_vars$`Chronic kidney diseases` <- 
    c(clinical_vars$`Chronic kidney diseases`, Cs(BMI, baseline_diabetes, blood_pressure_meds, systolic_bp, baseline_cvd))

clinical_vars$`Osteoarthritis` <- 
    c(clinical_vars$`Osteoarthritis`, Cs(BMI, strenuous_sports_4wks, heavy_DIY_4wks, heel_bone_mineral_density))

clinical_vars$`Lymphoma` <- 
    c(clinical_vars$`Osteoarthritis`, Cs(BMI, leukocyte_count))

clinical_vars$`All-cause mortality` <- 
    c(clinical_vars$`All-cause mortality`, Cs(overall_health, usual_walking_pace, baseline_diabetes, baseline_cvd, cancer_diagnosis, blood_pressure_meds, cholesterol_meds, baseline_cerebro, attendance_allowance, blue_badge, disability_allowance, BMI, systolic_bp, baseline_dyslipidemia, MDD))


# make sure unique
clinical_vars <- lapply(clinical_vars, function(x) {unique(x)})

```

```{r save dataset}

# save for quicker loading next time
save(all_data_ncd,
     scotwales_ncd,
     exposures_list,
     PRS_exposures,
     clinical_vars,
     file = paste0(path, "/R2_model_datasets_nov_09_2022.RData"))

```

## Run age and sex models

```{r run_models_age_sex}

# make index of diseases with at least 1 significant exposure
sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)

# list to store model results
R2_results_as <- as.list(seq(1, length(disease_list)))
names(R2_results_as) <- disease_list

# list to store sample sizes
sample_sizes <- as.list(seq(1, length(disease_list)))
names(sample_sizes) <- disease_list

for (k in seq_along(R2_results_as)) {
  
    # subset data
    # data <- all_data_ncd
    data <- all_data_ncd
    data <- lapply(data, identify_prev_cases_list[[k]])
    
    # make sub-list for model results
    R2_results_as[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {

        R2_results_as[[k]][[j]] <-
            coxph(as.formula(
                surv_list_age_sex[[k]]
            ), data = data[[j]])

    }
    
    # record sample size
    rm(data)
    
    cat("\r", paste("completed:", paste(k, length(R2_results_as), sep = "/")))
}


# subset to only models that were run
exclude <- which(names(exposures_list) == "Brain cancer")
R2_results_as <- R2_results_as[-exclude]

# create table_as for results
table_as <- data.frame(matrix(ncol = 5, nrow = length(R2_results_as)))
colnames(table_as) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# get R2 and C-index for all models
for (k in seq_along(R2_results_as)) {

    # get r-squared across imputations
    avg_rsq <- c(coxr2(R2_results_as[[k]][[1]])$rsq,
                 coxr2(R2_results_as[[k]][[2]])$rsq,
                 coxr2(R2_results_as[[k]][[3]])$rsq,
                 coxr2(R2_results_as[[k]][[4]])$rsq,
                 coxr2(R2_results_as[[k]][[5]])$rsq)
    
    # transform to correlation scale (sqrt)
    avg_rsq <- sapply(avg_rsq, sqrt)
    
    # transform using Fischer's Z
    avg_rsq <- sapply(avg_rsq, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_rsq <- mean(avg_rsq)
    
    # transform coefficients back to r scale
    avg_rsq <- FisherZInv(avg_rsq)
    
    # transform r squared scale
    avg_rsq <- (avg_rsq)^2
    
    # add to table_as
    table_as[k,3] <- avg_rsq
    
    
    # get c-index across imputations
    avg_index <- c(summary(R2_results_as[[k]][[1]])$concordance[1],
                   summary(R2_results_as[[k]][[2]])$concordance[1],
                   summary(R2_results_as[[k]][[3]])$concordance[1],
                   summary(R2_results_as[[k]][[4]])$concordance[1],
                   summary(R2_results_as[[k]][[5]])$concordance[1])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table_as
    table_as[k,2] <- avg_index


}

table_as$Disease <- names(R2_results_as)

# add in number of cases and sample sizes
for (k in seq_along(R2_results_as)) {

    table_as[k,4] <- R2_results_as[[k]][[1]]$nevent
    table_as[k,5] <- R2_results_as[[k]][[1]]$n
}


### save
save(R2_results_as,
     table_as,
     file = paste0(path, "/results/incident disease/R2_models_aug_25_2022_age_sex.RData"))

```

## Validate age and sex models in Scotland/Wales dataset

```{r prediction_age_sex, eval=TRUE}

# subset surv list
indicator_val <- indicator[sig_exposures]
surv_time_val <- surv_time[sig_exposures]

# subset prev cases function list
identify_prev_cases_list_val <- identify_prev_cases_list[sig_exposures]

# get prediction values for each disease
# pred_list <- as.list(seq_along(R2_results_exp))
# cindex_list <- as.list(seq_along(R2_results_exp))

pred_list <- as.list(seq_along(R2_results_as))
cindex_list_as <- as.list(seq_along(R2_results_as))
rsq_list_as <- as.list(seq_along(R2_results_as))

table_scotwales_as <- data.frame(matrix(ncol = 5, nrow = length(cindex_list_as)))
colnames(table_scotwales_as) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")


for (k in seq_along(pred_list)) {
    
    # subset data
    data <- scotwales_ncd
    data <- lapply(data, identify_prev_cases_list_val[[k]])
    
    # make sub-list for model results
    pred_list[[k]] <- as.list(seq(1, length(data)))
    cindex_list_as[[k]] <- as.list(seq(1, length(data)))
    rsq_list_as[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        # get linear predictors using scotwales data
        pred_list[[k]][[j]] <-
            predict(R2_results_as[[k]][[j]],
                    newdata = data[[j]])
        
        # Surv object in scotwales
        yrep <- Surv(data[[j]][[surv_time_val[k]]], data[[j]][[indicator_val[k]]])
        # calculate concordance outcomes using linear predictors and Surv object
        rs.corr <- rcorr.cens(x = pred_list[[k]][[j]], S = yrep)
        # take Dxy, negate, and then get C using formula Dxy=2*(C0.5)
        cindex_list_as[[k]][[j]] <- (-rs.corr[2]/2) + 0.5
        
        # Get R-squared using predicted values from scotwales data
        cox <- coxph(yrep ~ pred_list[[k]][[j]], data = data[[j]])
        rsq_list_as[[k]][[j]] <- coxr2(cox)
        
    }
    
    table_scotwales_as[k,4] <- as.numeric(table(data[[1]][indicator_val[k]])[2])
    table_scotwales_as[k,5] <- nrow(data[[1]])
    
    # record sample size
    rm(data)
    
    cat("\r", paste("completed:", paste(k, length(pred_list), sep = "/")))
    
}

### For further discussion of Dxy calculation, see comment here from Harrell on using linear predictors with rcorr.cens: https://stats.stackexchange.com/questions/48298/computing-c-index-for-an-external-validation-of-a-cox-ph-model-with-r#comment93976_48369


# get C-index for all models
for (k in seq_along(cindex_list_as)) {
    
    # Rsq
    avg_index <- c(rsq_list_as[[k]][[1]]$rsq,
                   rsq_list_as[[k]][[2]]$rsq,
                   rsq_list_as[[k]][[3]]$rsq,
                   rsq_list_as[[k]][[4]]$rsq,
                   rsq_list_as[[k]][[5]]$rsq)
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_as[k,3] <- avg_index
    
    # C-index
    avg_index <- c(cindex_list_as[[k]][[1]],
                   cindex_list_as[[k]][[2]],
                   cindex_list_as[[k]][[3]],
                   cindex_list_as[[k]][[4]],
                   cindex_list_as[[k]][[5]])
    
   
### save # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_as[k,2] <- avg_index

}

table_scotwales_as$Disease <- names(R2_results_as)

save(rsq_list_as,
     cindex_list_as,
     table_scotwales_as,
     file = paste0(path, "/results/incident disease/R2_models_aug_25_2022_scotwales_age_sex.RData"))

```

## Run exposome models

```{r run_models_exposome}

# make index of diseases with at least 1 significant exposure
sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)

# make list of diseases with at least 1 significant exposure
sig_diseases <- disease_list[sig_exposures]

# list to store model results
R2_results_exp <- as.list(seq(1, length(disease_list)))
names(R2_results_exp) <- disease_list

# list to store sample sizes
sample_sizes <- as.list(seq(1, length(disease_list)))
names(sample_sizes) <- disease_list

# make model exposures
model_exposures <- lapply(exposures_list, paste, collapse = " + ")
cores <- NULL
# run all models
for (k in sig_exposures) {
  
    # subset data
    data <- all_data_ncd
    data <- lapply(data, identify_prev_cases_list[[k]])
    
    
    # make sub-list for model results
    R2_results_exp[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {

        R2_results_exp[[k]][[j]] <-
            coxph(as.formula(
                paste0(
                    surv_list[[k]],
                    model_exposures[[k]])
            ), data = data[[j]])
    }
    
    # remove
    rm(data)
    invisible(gc())
    
    cat("\r", paste("completed:", paste(k, length(R2_results_exp), sep = "/")))
}


# subset to only models that were run
exclude <- which(sapply(R2_results_exp, is.numeric))
R2_results_exp <- R2_results_exp[-exclude]

# create table_exp for results
table_exp <- data.frame(matrix(ncol = 5, nrow = length(R2_results_exp)))
colnames(table_exp) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# get R2 and C-index for all models
for (k in seq_along(R2_results_exp)) {

    # get r-squared across imputations
    avg_rsq <- c(coxr2(R2_results_exp[[k]][[1]])$rsq,
                 coxr2(R2_results_exp[[k]][[2]])$rsq,
                 coxr2(R2_results_exp[[k]][[3]])$rsq,
                 coxr2(R2_results_exp[[k]][[4]])$rsq,
                 coxr2(R2_results_exp[[k]][[5]])$rsq)
    
    # transform to correlation scale (sqrt)
    avg_rsq <- sapply(avg_rsq, sqrt)
    
    # transform using Fischer's Z
    avg_rsq <- sapply(avg_rsq, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_rsq <- mean(avg_rsq)
    
    # transform coefficients back to r scale
    avg_rsq <- FisherZInv(avg_rsq)
    
    # transform r squared scale
    avg_rsq <- (avg_rsq)^2
    
    # add to table_exp
    table_exp[k,3] <- avg_rsq
    
    
    # get c-index across imputations
    avg_index <- c(summary(R2_results_exp[[k]][[1]])$concordance[1],
                   summary(R2_results_exp[[k]][[2]])$concordance[1],
                   summary(R2_results_exp[[k]][[3]])$concordance[1],
                   summary(R2_results_exp[[k]][[4]])$concordance[1],
                   summary(R2_results_exp[[k]][[5]])$concordance[1])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table_exp
    table_exp[k,2] <- avg_index


}

table_exp$Disease <- names(R2_results_exp)

# add in number of cases and sample sizes
for (k in seq_along(R2_results_exp)) {

    table_exp[k,4] <- R2_results_exp[[k]][[1]]$nevent
    table_exp[k,5] <- R2_results_exp[[k]][[1]]$n
}

### save
save(R2_results_exp,
     table_exp,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_exposome.RData"))


```

## Check proportional hazards in exposome models

```{r ph_check_exp}


sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)
cph_identify_prev <- identify_prev_cases_list[sig_exposures]
surv_list <- surv_list[sig_exposures]
model_exposures <- lapply(exposures_list, paste, collapse = " + ")
model_exposures <- model_exposures[sig_exposures]

j <- 1

# run check in first imputed dataset of each model result
test <- as.list(seq_along(cph_identify_prev))
for (k in 1:length(cph_identify_prev)) {
    data <- lapply(all_data_ncd, cph_identify_prev[[k]])
    test[[k]] <- cox.zph(R2_results_exp[[k]][[j]])
    rm(data)
    cat("\r", paste("completed:", paste(k, length(R2_results_exp), sep = "/")))
}

names(test) <- disease_list[sig_exposures]

# save
save(test,
     file = paste0(path, "/results/incident disease/cph_check_nov_09_2022_exposome.RData"))

# return to full surv_list
surv_list <- surv_list_copy
```

## check residual plots - exposome models

```{r ph_residuals_exposome}

load(paste0(path, "/results/incident disease/cph_check_nov_09_2022_exposome.RData"))

### find variables with low p-value from coxzph test
cph_var <- sapply(test, function(x){rownames(as.data.frame(x$table))[which(as.data.frame(x$table)$p < 0.00001)]})
cph_var <- lapply(cph_var, function(x) {x <- x[x != "GLOBAL"]})


### check schoenfield residuals for each

# residuals **could be skewed - test further
plot(test$`All-cause mortality`, var = cph_var$`All-cause mortality`, resid = TRUE)

# residuals look fine
for (k in 1:length(cph_var$`Prostate cancer`)) {
plot(test$`Prostate cancer`, var = cph_var$`Prostate cancer`[k], resid = TRUE)
}

# residuals look fine
for (k in 1:length(cph_var$`Ischemic heart disease`)) {
plot(test$`Ischemic heart disease`, var = cph_var$`Ischemic heart disease`[k], resid = TRUE)
}

# residuals look fine
for (k in 1:length(cph_var$`Emphysema, COPD`)) {
plot(test$`Emphysema, COPD`, var = cph_var$`Emphysema, COPD`[k], resid = TRUE)
}

# residuals look fine
plot(test$`Chronic liver diseases`, var = cph_var$`Chronic liver diseases`, resid = TRUE)

# residuals look fine
for (k in 1:length(cph_var$`Chronic kidney diseases`)) {
plot(test$`Chronic kidney diseases`, var = cph_var$`Chronic kidney diseases`[k], resid = TRUE)
}

# residuals look fine
plot(test$`Rheumatoid arthritis`, var = cph_var$`Rheumatoid arthritis`, resid = TRUE)

# residuals look fine
for (k in 1:length(cph_var$Osteoarthritis)) {
plot(test$Osteoarthritis, var = cph_var$Osteoarthritis[k], resid = TRUE)
}

# one variable to correct:
## all-cause mortality

```

## Re-run exposome models with time interaction

```{r ph_run_models_exposome}

# get list of outcomes to redo with time interaction
redo <- which(names(cph_var) %in% c("All-cause mortality"))

# make index of diseases with at least 1 significant exposure
sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)
cph_exposures_list <- exposures_list[sig_exposures]
cph_identify_prev <- identify_prev_cases_list[sig_exposures]
surv_list_time <- surv_list_tt[sig_exposures]
cph_indicator <- indicator[sig_exposures]
cph_surv <- surv_time[sig_exposures]
ph.test <- as.list(seq(1, length(R2_results_exp)))

# make model exposures
time_int_var <- cph_var
for (k in redo) {
    time_int_var[[k]] <- paste0(time_int_var[[k]], ":Start_time")
}

for (k in seq_along(time_int_var)[-redo]) {
    time_int_var[[k]] <- vector()
}

time_int_var <- lapply(time_int_var, function(x) paste(x, collapse = " + "))
model_exposures <- lapply(cph_exposures_list, paste, collapse = " + ")


library(magrittr)
library(Greg)

# copy old R2 results
R2_results_exp_cph <- R2_results_exp

# redo interaction models
for (k in redo) {
  
    # subset data
    data <- all_data_ncd
    data <- lapply(data, cph_identify_prev[[k]])
    
    # timesplit function
    data <- lapply(data, function(x){
        # reduce data to just colnames of model variables to improve runtime with timesplit
        x <- x[c(cph_exposures_list[[k]], cph_indicator[k], cph_surv[k], "recruitment_age", "age_scaled", "sex")]
        # turn survival time to years to be on same scale as age
        x$time <- as.numeric(x[[cph_surv[k]]] / 365.25)
        # timesplit data
        x <- x %>% 
            timeSplitter(by = 2,  
                         event_var = cph_indicator[k],
                         event_start_status = 0,
                         time_var = "time",
                         time_related_vars = "recruitment_age")
        return(x)
    })
    
    # make sub-list for model results
    R2_results_exp_cph[[k]] <- as.list(seq(1, length(data)))
    ph.test[[k]] <- as.list(seq(1, length(data)))
    # run model in each imputed dataset
    for (j in seq_along(data)) {

        R2_results_exp_cph[[k]][[j]] <-
            coxph(as.formula(
                paste(
                    paste0(
                        surv_list_time[[k]],
                        model_exposures[[k]]),
                    time_int_var[[k]], sep = " + ")
                ), data = data[[j]])
        
        # test ph assumption, stop if not fixed
        ph.test[[k]][[j]] <- cox.zph(R2_results_exp_cph[[k]][[j]])
        # trial <- any(as.data.frame(ph.test$table)$p < 0.00001)
        # if (trial == TRUE) {
            # print(ph.test)
            # stop("non-proportional hazards not fixed")
        # }
    }
    
    # remove
    rm(data)
    invisible(gc())
    
    cat("\r", paste("completed:", paste(k, length(R2_results_exp_cph), sep = "/")))
}

# create table_exp_cph for results
table_exp_cph <- data.frame(matrix(ncol = 5, nrow = length(R2_results_exp_cph)))
colnames(table_exp_cph) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# get R2 and C-index for all models
for (k in seq_along(R2_results_exp_cph)) {

    # get r-squared across imputations
    avg_rsq <- c(coxr2(R2_results_exp_cph[[k]][[1]])$rsq,
                 coxr2(R2_results_exp_cph[[k]][[2]])$rsq,
                 coxr2(R2_results_exp_cph[[k]][[3]])$rsq,
                 coxr2(R2_results_exp_cph[[k]][[4]])$rsq,
                 coxr2(R2_results_exp_cph[[k]][[5]])$rsq)
    
    # transform to correlation scale (sqrt)
    avg_rsq <- sapply(avg_rsq, sqrt)
    
    # transform using Fischer's Z
    avg_rsq <- sapply(avg_rsq, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_rsq <- mean(avg_rsq)
    
    # transform coefficients back to r scale
    avg_rsq <- FisherZInv(avg_rsq)
    
    # transform r squared scale
    avg_rsq <- (avg_rsq)^2
    
    # add to table_exp_cph
    table_exp_cph[k,3] <- avg_rsq
    
    
    # get c-index across imputations
    avg_index <- c(summary(R2_results_exp_cph[[k]][[1]])$concordance[1],
                   summary(R2_results_exp_cph[[k]][[2]])$concordance[1],
                   summary(R2_results_exp_cph[[k]][[3]])$concordance[1],
                   summary(R2_results_exp_cph[[k]][[4]])$concordance[1],
                   summary(R2_results_exp_cph[[k]][[5]])$concordance[1])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table_exp_cph
    table_exp_cph[k,2] <- avg_index


}

table_exp_cph$Disease <- names(R2_results_exp_cph)

# add in number of cases and sample sizes
for (k in seq_along(R2_results_exp_cph)[-redo]) {

    table_exp_cph[k,4] <- R2_results_exp_cph[[k]][[1]]$nevent
    table_exp_cph[k,5] <- R2_results_exp_cph[[k]][[1]]$n
}

# use old table for new time split models, since sample will be in millions
for (k in seq_along(R2_results_exp_cph)[redo]) {

    table_exp_cph[k,4] <- table_exp[k,4]
    table_exp_cph[k,5] <- table_exp[k,5]
}


### save
save(R2_results_exp_cph,
     ph.test,
     table_exp_cph,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_exposome_cph.RData"))

```

## Validate exposome model in Scotland/Wales dataset

```{r prediction_exposome, eval=TRUE}

# subset surv list
indicator_val <- indicator[sig_exposures]
surv_time_val <- surv_time[sig_exposures]

# subset prev cases function list
identify_prev_cases_list_val <- identify_prev_cases_list[sig_exposures]

# get prediction values for each disease
pred_list <- as.list(seq_along(R2_results_exp))
cindex_list_exp <- as.list(seq_along(R2_results_exp))
rsq_list_exp <- as.list(seq_along(R2_results_exp))

# create table for results
table_scotwales_exp <- data.frame(matrix(ncol = 5, nrow = length(cindex_list_exp)))
colnames(table_scotwales_exp) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# run
for (k in seq_along(pred_list)) {
    
    # subset data
    data <- scotwales_ncd
    data <- lapply(data, identify_prev_cases_list_val[[k]])
    
    # make sub-list for model results
    pred_list[[k]] <- as.list(seq(1, length(data)))
    cindex_list_exp[[k]] <- as.list(seq(1, length(data)))
    rsq_list_exp[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        # get linear predictors using scotwales data
        pred_list[[k]][[j]] <-
            predict(R2_results_exp[[k]][[j]],
                    newdata = data[[j]])
        
        # Surv object in scotwales
        yrep <- Surv(data[[j]][[surv_time_val[k]]], data[[j]][[indicator_val[k]]])
        # calculate concordance outcomes using linear predictors and Surv object
        rs.corr <- rcorr.cens(x = pred_list[[k]][[j]], S = yrep)
        # take Dxy, negate, and then get C using formula Dxy=2*(C0.5)
        cindex_list_exp[[k]][[j]] <- (-rs.corr[2]/2) + 0.5
        
        # Get R-squared using predicted values from scotwales data
        cox <- coxph(yrep ~ pred_list[[k]][[j]], data = data[[j]])
        rsq_list_exp[[k]][[j]] <- coxr2(cox)
        
    }
    
    table_scotwales_exp[k,4] <- as.numeric(table(data[[1]][indicator_val[k]])[2])
    table_scotwales_exp[k,5] <- nrow(data[[1]])
    
    # record sample size
    rm(data)
    
    cat("\r", paste("completed:", paste(k, length(pred_list), sep = "/")))
    
}

### For further discussion of Dxy calculation, see comment here from Harrell on using linear predictors with rcorr.cens: https://stats.stackexchange.com/questions/48298/computing-c-index-for-an-external-validation-of-a-cox-ph-model-with-r#comment93976_48369


# get C-index for all models
for (k in seq_along(cindex_list_exp)) {
    
    # Rsq
    avg_index <- c(rsq_list_exp[[k]][[1]]$rsq,
                   rsq_list_exp[[k]][[2]]$rsq,
                   rsq_list_exp[[k]][[3]]$rsq,
                   rsq_list_exp[[k]][[4]]$rsq,
                   rsq_list_exp[[k]][[5]]$rsq)
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_exp[k,3] <- avg_index
    
    # C-index
    avg_index <- c(cindex_list_exp[[k]][[1]],
                   cindex_list_exp[[k]][[2]],
                   cindex_list_exp[[k]][[3]],
                   cindex_list_exp[[k]][[4]],
                   cindex_list_exp[[k]][[5]])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_exp[k,2] <- avg_index

}

table_scotwales_exp$Disease <- names(R2_results_exp)

### save
save(rsq_list_exp,
     cindex_list_exp,
     table_scotwales_exp,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_exposome.RData"))

```

## Validate exposome model in Scotland/Wales dataset - adding time interaction results

```{r prediction_exposome_cph, eval=TRUE}

# subset surv list
indicator_val <- indicator[sig_exposures]
surv_time_val <- surv_time[sig_exposures]

# subset prev cases function list
identify_prev_cases_list_val <- identify_prev_cases_list[sig_exposures]

# list for prediction values for each disease
pred_list <- as.list(seq_along(R2_results_exp_cph))

# copy results from previous
cindex_list_exp_cph <- cindex_list_exp
rsq_list_exp_cph <- rsq_list_exp

table_scotwales_exp_cph <- table_scotwales_exp

for (k in redo) {
    
    # subset data
    data <- scotwales_ncd
    data <- lapply(data, identify_prev_cases_list_val[[k]])
    data2 <- data
    
    data <- lapply(data, function(x){
        # reduce data to just colnames of model variables to improve runtime with timesplit
        x <- x[c(cph_exposures_list[[k]], cph_indicator[k], cph_surv[k], "recruitment_age", "age_scaled", "sex")]
        # turn survival time to years to be on same scale as age
        x$time <- as.numeric(x[[cph_surv[k]]] / 365.25)
        # timesplit data
        x <- x %>% 
            timeSplitter(by = 2,  
                         event_var = cph_indicator[k],
                         event_start_status = 0,
                         time_var = "time",
                         time_related_vars = "recruitment_age")
        return(x)
    })
    
    # make sub-list for model results
    pred_list[[k]] <- as.list(seq(1, length(data)))
    cindex_list_exp_cph[[k]] <- as.list(seq(1, length(data)))
    rsq_list_exp_cph[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        # get linear predictors using scotwales data
        pred_list[[k]][[j]] <-
            predict(R2_results_exp_cph[[k]][[j]],
                    newdata = data[[j]])
        # Surv object in scotwales using interval time for cph interaction
        yrep <- Surv(data[[j]][["Start_time"]], data[[j]][["Stop_time"]], data[[j]][[indicator_val[k]]])
        # Get R-squared using predicted values from scotwales data
        cox <- coxph(yrep ~ pred_list[[k]][[j]], data = data[[j]])
        cindex_list_exp_cph[[k]][[j]] <- summary(cox)$concordance[1]
        rsq_list_exp_cph[[k]][[j]] <- coxr2(cox)
    }
    
    # add no. cases and sample size to table
    table_scotwales_exp_cph[k,4] <- as.numeric(table(data2[[1]][indicator_val[k]])[2])
    table_scotwales_exp_cph[k,5] <- nrow(data2[[1]])
    
    # record sample size
    rm(data, data2)
    
    cat("\r", paste("completed:", paste(k, length(pred_list), sep = "/")))
    
}

### For further discussion of Dxy calculation, see comment here from Harrell on using linear predictors with rcorr.cens: https://stats.stackexchange.com/questions/48298/computing-c-index-for-an-external-validation-of-a-cox-ph-model-with-r#comment93976_48369


# get C-index for all models
for (k in redo) {
    
    # Rsq
    avg_index <- c(rsq_list_exp_cph[[k]][[1]]$rsq,
                   rsq_list_exp_cph[[k]][[2]]$rsq,
                   rsq_list_exp_cph[[k]][[3]]$rsq,
                   rsq_list_exp_cph[[k]][[4]]$rsq,
                   rsq_list_exp_cph[[k]][[5]]$rsq)
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_exp_cph[k,3] <- avg_index
    
    # C-index
    avg_index <- c(cindex_list_exp_cph[[k]][[1]],
                   cindex_list_exp_cph[[k]][[2]],
                   cindex_list_exp_cph[[k]][[3]],
                   cindex_list_exp_cph[[k]][[4]],
                   cindex_list_exp_cph[[k]][[5]])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_exp_cph[k,2] <- avg_index

}

### save
save(rsq_list_exp_cph,
     cindex_list_exp_cph,
     table_scotwales_exp_cph,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_exposome_cph.RData"))

```

## Run exposome + PRS models

```{r run_models_prs}

# subset to diseases with PRS
PRS_index <- which(disease_list %in% PRS_disease_list)
PRS_identify_prev <- identify_prev_cases_list[PRS_index]
PRS_surv_list <- surv_list[PRS_index]

# list to store model results
R2_results_prs <- as.list(seq(1, length(PRS_disease_list)))
names(R2_results_prs) <- PRS_disease_list

# list to store sample sizes
sample_sizes <- as.list(seq(1, length(PRS_disease_list)))
names(sample_sizes) <- PRS_disease_list

# make model exposures
model_exposures <- lapply(PRS_exposures, paste, collapse = " + ")

# run all models
for (k in seq_along(R2_results_prs)) {
  
    # subset data
    data <- all_data_ncd
    data <- lapply(data, PRS_identify_prev[[k]])
    
    # make sub-list for model results
    R2_results_prs[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {

        R2_results_prs[[k]][[j]] <-
            coxph(as.formula(
                paste0(
                    PRS_surv_list[[k]],
                    model_exposures[[k]])
            ), data = data[[j]])
    }
    
    # record sample size
    rm(data)
    
    cat("\r", paste("completed:", paste(k, length(R2_results_prs), sep = "/")))
}

# create table_prs for results
table_prs <- data.frame(matrix(ncol = 5, nrow = length(R2_results_prs)))
colnames(table_prs) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# get R2 and C-index for all models
for (k in seq_along(R2_results_prs)) {

    # get r-squared across imputations
    avg_rsq <- c(coxr2(R2_results_prs[[k]][[1]])$rsq,
                 coxr2(R2_results_prs[[k]][[2]])$rsq,
                 coxr2(R2_results_prs[[k]][[3]])$rsq,
                 coxr2(R2_results_prs[[k]][[4]])$rsq,
                 coxr2(R2_results_prs[[k]][[5]])$rsq)
    
    # transform to correlation scale (sqrt)
    avg_rsq <- sapply(avg_rsq, sqrt)
    
    # transform using Fischer's Z
    avg_rsq <- sapply(avg_rsq, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_rsq <- mean(avg_rsq)
    
    # transform coefficients back to r scale
    avg_rsq <- FisherZInv(avg_rsq)
    
    # transform r squared scale
    avg_rsq <- (avg_rsq)^2
    
    # add to table_prs
    table_prs[k,3] <- avg_rsq
    
    
    # get c-index across imputations
    avg_index <- c(summary(R2_results_prs[[k]][[1]])$concordance[1],
                   summary(R2_results_prs[[k]][[2]])$concordance[1],
                   summary(R2_results_prs[[k]][[3]])$concordance[1],
                   summary(R2_results_prs[[k]][[4]])$concordance[1],
                   summary(R2_results_prs[[k]][[5]])$concordance[1])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table_prs
    table_prs[k,2] <- avg_index


}

table_prs$Disease <- names(R2_results_prs)

# add in number of cases and sample sizes
for (k in seq_along(R2_results_prs)) {

    table_prs[k,4] <- R2_results_prs[[k]][[1]]$nevent
    table_prs[k,5] <- R2_results_prs[[k]][[1]]$n
}


### save
save(R2_results_prs,
     table_prs,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_PRS.RData"))

```

## Check proportional hazards for exposome + PRS models

```{r ph_check_prs}

PRS_index <- which(disease_list %in% PRS_disease_list)

cph_identify_prev <- identify_prev_cases_list[PRS_index]
cph_surv_list <- surv_list[PRS_index]
model_exposures <- lapply(PRS_exposures, paste, collapse = " + ")

j <- 1

# run
test <- as.list(seq_along(cph_identify_prev))
for (k in 1:length(cph_identify_prev)) {
    data <- lapply(all_data_ncd, cph_identify_prev[[k]])
    test[[k]] <- cox.zph(R2_results_prs[[k]][[j]])
    rm(data)
    cat("\r", paste("completed:", paste(k, length(R2_results_prs), sep = "/")))
}

names(test) <- disease_list[PRS_index]

# save
save(test,
     file = paste0(path, "/results/incident disease/cph_check_nov_09_2022_prs.RData"))
```

## Check residuals for exposome + PRS models

```{r ph_residuals_prs}

### find variables with low p-value from coxzph test
cph_var <- sapply(test, function(x){rownames(as.data.frame(x$table))[which(as.data.frame(x$table)$p < 0.00001)]})
cph_var <- lapply(cph_var, function(x) {x <- x[x != "GLOBAL"]})


### check schoenfield residuals for each

# residuals **could be skewed - test further
plot(test$`All-cause mortality`, var = cph_var$`All-cause mortality`, resid = TRUE)

# residuals look fine
for (k in 1:length(cph_var$`Prostate cancer`)) {
plot(test$`Prostate cancer`, var = cph_var$`Prostate cancer`[k], resid = TRUE)
}

# residuals look fine
for (k in 1:length(cph_var$`Ischemic heart disease`)) {
plot(test$`Ischemic heart disease`, var = cph_var$`Ischemic heart disease`[k], resid = TRUE)
}

# residuals look fine
plot(test$`Rheumatoid arthritis`, var = cph_var$`Rheumatoid arthritis`, resid = TRUE)

# one variables to correct:
## all-cause mortality

```

## Re-run exposome + PRS models with time interaction

```{r ph_run_models_prs}

indicator <- c(
    Cs(
        ACM_event_indicator,
        colon_cancer_event,
        lung_cancer_event,
        eso_cancer_event,
        liver_cancer_event,
        pancreatic_cancer_event,
        brain_cancer_event,
        leukemia_event,
        lymphoma_event,
        breast_cancer_event,
        ovarian_cancer_event,
        prostate_cancer_event,
        diabetes_event,
        CVD_event,
        cerebro_event,
        COPD_event,
        liver_event,
        kidney_event,
        all_dementia_event,
        vasc_dementia_event,
        alzheimers_event,
        parkinsons_event,
        rheumatoid_event,
        macular_event,
        osteoporosis_event,
        osteoarthritis_event
    )
)

surv_time <- c(
    Cs(
        ACM_survival_time,
        colon_cancer_survival_time,
        lung_cancer_survival_time,
        eso_cancer_survival_time,
        liver_cancer_survival_time,
        pancreatic_cancer_survival_time,
        brain_cancer_survival_time,
        leukemia_survival_time,
        lymphoma_survival_time,
        breast_cancer_survival_time,
        ovarian_cancer_survival_time,
        prostate_cancer_survival_time,
        diabetes_survival_time,
        CVD_survival_time,
        cerebro_survival_time,
        COPD_survival_time,
        liver_survival_time,
        kidney_survival_time,
        all_dementia_survival_time,
        vasc_dementia_survival_time,
        alzheimers_survival_time,
        parkinsons_survival_time,
        rheumatoid_survival_time,
        macular_survival_time,
        osteoporosis_survival_time,
        osteoarthritis_survival_time
    )
)

# get index of outcomes to redo with time interaction
redo <- which(names(R2_results_prs) %in% c("All-cause mortality"))

# make index of diseases with PRS
PRS_index <- which(disease_list %in% PRS_disease_list)
cph_exposures_list <- PRS_exposures
cph_identify_prev <- identify_prev_cases_list[PRS_index]
surv_list_time <- surv_list_tt[PRS_index]
cph_indicator <- indicator[PRS_index]
cph_surv <- surv_time[PRS_index]
ph.test <- as.list(seq(1, length(R2_results_prs)))

# make model exposures
time_int_var <- cph_var
for (k in redo) {
    time_int_var[[k]] <- paste0(time_int_var[[k]], ":Start_time")
}

for (k in seq_along(time_int_var)[-redo]) {
    time_int_var[[k]] <- vector()
}

time_int_var <- lapply(time_int_var, function(x) paste(x, collapse = " + "))
model_exposures <- lapply(cph_exposures_list, paste, collapse = " + ")


library(magrittr)
library(Greg)

# copy old R2 results
R2_results_prs_cph <- R2_results_prs

# redo interaction models
for (k in redo) {
  
    # subset data
    data <- all_data_ncd
    data <- lapply(data, cph_identify_prev[[k]])
    
    # timesplit function
    data <- lapply(data, function(x){
        # reduce data to just colnames of model variables to improve runtime with timesplit
        x <- x[c(cph_exposures_list[[k]], cph_indicator[k], cph_surv[k], "recruitment_age", "age_scaled", "sex")]
        # turn survival time to years to be on same scale as age
        x$time <- as.numeric(x[[cph_surv[k]]] / 365.25)
        # timesplit data
        x <- x %>% 
            timeSplitter(by = 2,  
                         event_var = cph_indicator[k],
                         event_start_status = 0,
                         time_var = "time",
                         time_related_vars = "recruitment_age")
        return(x)
    })
    
    # make sub-list for model results
    R2_results_prs_cph[[k]] <- as.list(seq(1, length(data)))
    ph.test[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {

        R2_results_prs_cph[[k]][[j]] <-
            coxph(as.formula(
                paste(
                    paste0(
                        surv_list_time[[k]],
                        model_exposures[[k]]),
                    time_int_var[[k]], sep = " + ")
                ), data = data[[j]])
        
        # test ph assumption, stop if not fixed
        ph.test[[k]][[j]] <- cox.zph(R2_results_prs_cph[[k]][[j]])
    }
    
    # remove
    rm(data)
    invisible(gc())
    
    cat("\r", paste("completed:", paste(k, length(R2_results_prs_cph), sep = "/")))
}


# create table_prs_cph for results
table_prs_cph <- data.frame(matrix(ncol = 5, nrow = length(R2_results_prs_cph)))
colnames(table_prs_cph) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# get R2 and C-index for all models
for (k in seq_along(R2_results_prs_cph)) {

    # get r-squared across imputations
    avg_rsq <- c(coxr2(R2_results_prs_cph[[k]][[1]])$rsq,
                 coxr2(R2_results_prs_cph[[k]][[2]])$rsq,
                 coxr2(R2_results_prs_cph[[k]][[3]])$rsq,
                 coxr2(R2_results_prs_cph[[k]][[4]])$rsq,
                 coxr2(R2_results_prs_cph[[k]][[5]])$rsq)
    
    # transform to correlation scale (sqrt)
    avg_rsq <- sapply(avg_rsq, sqrt)
    
    # transform using Fischer's Z
    avg_rsq <- sapply(avg_rsq, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_rsq <- mean(avg_rsq)
    
    # transform coefficients back to r scale
    avg_rsq <- FisherZInv(avg_rsq)
    
    # transform r squared scale
    avg_rsq <- (avg_rsq)^2
    
    # add to table_prs_cph
    table_prs_cph[k,3] <- avg_rsq
    
    
    # get c-index across imputations
    avg_index <- c(summary(R2_results_prs_cph[[k]][[1]])$concordance[1],
                   summary(R2_results_prs_cph[[k]][[2]])$concordance[1],
                   summary(R2_results_prs_cph[[k]][[3]])$concordance[1],
                   summary(R2_results_prs_cph[[k]][[4]])$concordance[1],
                   summary(R2_results_prs_cph[[k]][[5]])$concordance[1])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table_prs_cph
    table_prs_cph[k,2] <- avg_index


}

table_prs_cph$Disease <- names(R2_results_prs_cph)

# add in number of cases and sample sizes
for (k in seq_along(R2_results_prs_cph)[-redo]) {

    table_prs_cph[k,4] <- R2_results_prs_cph[[k]][[1]]$nevent
    table_prs_cph[k,5] <- R2_results_prs_cph[[k]][[1]]$n
}

# use old table for new time split models, since sample will be in millions
for (k in seq_along(R2_results_prs_cph)[redo]) {

    table_prs_cph[k,4] <- table_prs[k,4]
    table_prs_cph[k,5] <- table_prs[k,5]
}


### save
save(R2_results_prs_cph,
     ph.test,
     table_prs_cph,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_PRS_cph.RData"))

```

## Validate exposome + PRS model in Scotland/Wales dataset

```{r prediction_prs, eval=TRUE}

PRS_index <- which(disease_list %in% PRS_disease_list)
PRS_identify_prev <- identify_prev_cases_list[PRS_index]
PRS_surv_list <- surv_list[PRS_index]

# subset surv list
indicator_val <- indicator[PRS_index]
surv_time_val <- surv_time[PRS_index]

pred_list <- as.list(seq_along(R2_results_prs))
cindex_list_prs <- as.list(seq_along(R2_results_prs))
rsq_list_prs <- as.list(seq_along(R2_results_prs))

table_scotwales_prs <- data.frame(matrix(ncol = 5, nrow = length(cindex_list_prs)))
colnames(table_scotwales_prs) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")


for (k in seq_along(pred_list)) {
    
    # subset data
    data <- scotwales_ncd
    data <- lapply(data, PRS_identify_prev[[k]])
    
    # make sub-list for model results
    pred_list[[k]] <- as.list(seq(1, length(data)))
    cindex_list_prs[[k]] <- as.list(seq(1, length(data)))
    rsq_list_prs[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        # get linear predictors using scotwales data
        pred_list[[k]][[j]] <-
            predict(R2_results_prs[[k]][[j]],
                    newdata = data[[j]])
        
        # Surv object in scotwales
        yrep <- Surv(data[[j]][[surv_time_val[k]]], data[[j]][[indicator_val[k]]])
        # calculate concordance outcomes using linear predictors and Surv object
        rs.corr <- rcorr.cens(x = pred_list[[k]][[j]], S = yrep)
        # take Dxy, negate, and then get C using formula Dxy=2*(C0.5)
        cindex_list_prs[[k]][[j]] <- (-rs.corr[2]/2) + 0.5
        
        # Get R-squared using predicted values from scotwales data
        cox <- coxph(yrep ~ pred_list[[k]][[j]], data = data[[j]])
        rsq_list_prs[[k]][[j]] <- coxr2(cox)
        
    }
    
    table_scotwales_prs[k,4] <- as.numeric(table(data[[1]][indicator_val[k]])[2])
    table_scotwales_prs[k,5] <- nrow(data[[1]])
    
    # record sample size
    rm(data)
    
    cat("\r", paste("completed:", paste(k, length(pred_list), sep = "/")))
    
}

### For further discussion of Dxy calculation, see comment here from Harrell on using linear predictors with rcorr.cens: https://stats.stackexchange.com/questions/48298/computing-c-index-for-an-external-validation-of-a-cox-ph-model-with-r#comment93976_48369


# get C-index for all models
for (k in seq_along(cindex_list_prs)) {
    
    # Rsq
    avg_index <- c(rsq_list_prs[[k]][[1]]$rsq,
                   rsq_list_prs[[k]][[2]]$rsq,
                   rsq_list_prs[[k]][[3]]$rsq,
                   rsq_list_prs[[k]][[4]]$rsq,
                   rsq_list_prs[[k]][[5]]$rsq)
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_prs[k,3] <- avg_index
    
    # C-index
    avg_index <- c(cindex_list_prs[[k]][[1]],
                   cindex_list_prs[[k]][[2]],
                   cindex_list_prs[[k]][[3]],
                   cindex_list_prs[[k]][[4]],
                   cindex_list_prs[[k]][[5]])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_prs[k,2] <- avg_index

}

table_scotwales_prs$Disease <- names(R2_results_prs)

### save
save(rsq_list_prs,
     cindex_list_prs,
     table_scotwales_prs,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_PRS.RData"))

```

## Validate exposome + PRS model in Scotland/Wales dataset - adding time interaction results

```{r prediction_prs_cph, eval=TRUE}

# get index of outcomes to redo with time interaction
redo <- which(names(R2_results_prs_cph) %in% c("All-cause mortality"))

# subset prev cases function list
PRS_index <- which(disease_list %in% PRS_disease_list)
identify_prev_cases_list_val <- identify_prev_cases_list[PRS_index]

# subset surv list
indicator_val <- indicator[PRS_index]
surv_time_val <- surv_time[PRS_index]

# list for prediction values for each disease
pred_list <- as.list(seq_along(R2_results_prs_cph))
cindex_list_prs_cph <- cindex_list_prs
rsq_list_prs_cph <- rsq_list_prs

table_scotwales_prs_cph <- table_scotwales_prs


for (k in redo) {
    
    # subset data
    data <- scotwales_ncd
    data <- lapply(data, identify_prev_cases_list_val[[k]])
    data2 <- data
    
    # timesplit function
    data <- lapply(data, function(x){
        # reduce data to just colnames of model variables to improve runtime with timesplit
        x <- x[c(PRS_exposures[[k]], indicator_val[k], cph_surv[k], "recruitment_age", "age_scaled", "sex")]
        # turn survival time to years to be on same scale as age
        x$time <- as.numeric(x[[cph_surv[k]]] / 365.25)
        # timesplit data
        x <- x %>% 
            timeSplitter(by = 2,  
                         event_var = indicator_val[k],
                         event_start_status = 0,
                         time_var = "time",
                         time_related_vars = "recruitment_age")
        return(x)
    })
    
    # make sub-list for model results
    pred_list[[k]] <- as.list(seq(1, length(data)))
    cindex_list_prs_cph[[k]] <- as.list(seq(1, length(data)))
    rsq_list_prs_cph[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        # get linear predictors using scotwales data
        pred_list[[k]][[j]] <-
            predict(R2_results_prs_cph[[k]][[j]],
                    newdata = data[[j]])
        
        # Surv object in scotwales using interval time for cph interaction
        yrep <- Surv(data[[j]][["Start_time"]], data[[j]][["Stop_time"]], data[[j]][[indicator_val[k]]])
        # Get R-squared using predicted values from scotwales data
        cox <- coxph(yrep ~ pred_list[[k]][[j]], data = data[[j]])
        cindex_list_prs_cph[[k]][[j]] <- summary(cox)$concordance[1]
        rsq_list_prs_cph[[k]][[j]] <- coxr2(cox)
    }
    
    
    table_scotwales_prs_cph[k,4] <- as.numeric(table(data2[[1]][indicator_val[k]])[2])
    table_scotwales_prs_cph[k,5] <- nrow(data2[[1]])
    
    # record sample size
    rm(data, data2)
    
    cat("\r", paste("completed:", paste(k, length(pred_list), sep = "/")))
    
}

### For further discussion of Dxy calculation, see comment here from Harrell on using linear predictors with rcorr.cens: https://stats.stackexchange.com/questions/48298/computing-c-index-for-an-external-validation-of-a-cox-ph-model-with-r#comment93976_48369


# get C-index for all models
for (k in seq_along(cindex_list_prs_cph)) {
    
    # Rsq
    avg_index <- c(rsq_list_prs_cph[[k]][[1]]$rsq,
                   rsq_list_prs_cph[[k]][[2]]$rsq,
                   rsq_list_prs_cph[[k]][[3]]$rsq,
                   rsq_list_prs_cph[[k]][[4]]$rsq,
                   rsq_list_prs_cph[[k]][[5]]$rsq)
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_prs_cph[k,3] <- avg_index
    
    # C-index
    avg_index <- c(cindex_list_prs_cph[[k]][[1]],
                   cindex_list_prs_cph[[k]][[2]],
                   cindex_list_prs_cph[[k]][[3]],
                   cindex_list_prs_cph[[k]][[4]],
                   cindex_list_prs_cph[[k]][[5]])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_prs_cph[k,2] <- avg_index

}

table_scotwales_prs_cph$Disease <- names(R2_results_prs_cph)

### save
save(rsq_list_prs_cph,
     cindex_list_prs_cph,
     table_scotwales_prs_cph,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_PRS_cph.RData"))

```

## Run exposome + PRS + clinical risk factor models

```{r run_models_clinical_factors, eval=TRUE}

# make index of diseases with at least 1 significant exposure
sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)

# make list of diseases with at least 1 significant exposure
sig_diseases <- disease_list[sig_exposures]

# list to store model results
R2_results_clin <- as.list(seq(1, length(disease_list)))
names(R2_results_clin) <- disease_list

# list to store sample sizes
sample_sizes <- as.list(seq(1, length(disease_list)))
names(sample_sizes) <- disease_list

# make model exposures
model_exposures <- lapply(clinical_vars, paste, collapse = " + ")

# run all models
for (k in sig_exposures) {
  
    # subset data
    data <- lapply(all_data_ncd, identify_prev_cases_list[[k]])
    
    # make sub-list for model results
    R2_results_clin[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        R2_results_clin[[k]][[j]] <-
            coxph(as.formula(
                paste0(
                    surv_list[[k]],
                    model_exposures[[k]])
            ), data = data[[j]])
        
    }
    
    # record sample size
    rm(data)
    
    cat("\r", paste("completed:", paste(k, length(R2_results_clin), sep = "/")))
}


# subset to only models that were run
exclude <- which(sapply(R2_results_clin, is.numeric))
R2_results_clin <- R2_results_clin[-exclude]

# create table_clin for results
table_clin <- data.frame(matrix(ncol = 5, nrow = length(R2_results_clin)))
colnames(table_clin) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# get R2 and C-index for all models
for (k in seq_along(R2_results_clin)) {

    # get r-squared across imputations
    avg_rsq <- c(coxr2(R2_results_clin[[k]][[1]])$rsq,
                 coxr2(R2_results_clin[[k]][[2]])$rsq,
                 coxr2(R2_results_clin[[k]][[3]])$rsq,
                 coxr2(R2_results_clin[[k]][[4]])$rsq,
                 coxr2(R2_results_clin[[k]][[5]])$rsq)
    
    # transform to correlation scale (sqrt)
    avg_rsq <- sapply(avg_rsq, sqrt)
    
    # transform using Fischer's Z
    avg_rsq <- sapply(avg_rsq, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_rsq <- mean(avg_rsq)
    
    # transform coefficients back to r scale
    avg_rsq <- FisherZInv(avg_rsq)
    
    # transform r squared scale
    avg_rsq <- (avg_rsq)^2
    
    # add to table_clin
    table_clin[k,3] <- avg_rsq
    
    
    # get c-index across imputations
    avg_index <- c(summary(R2_results_clin[[k]][[1]])$concordance[1],
                   summary(R2_results_clin[[k]][[2]])$concordance[1],
                   summary(R2_results_clin[[k]][[3]])$concordance[1],
                   summary(R2_results_clin[[k]][[4]])$concordance[1],
                   summary(R2_results_clin[[k]][[5]])$concordance[1])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table_clin
    table_clin[k,2] <- avg_index


}

table_clin$Disease <- names(R2_results_clin)

# add in number of cases and sample sizes
for (k in seq_along(R2_results_clin)) {

    table_clin[k,4] <- R2_results_clin[[k]][[1]]$nevent
    table_clin[k,5] <- R2_results_clin[[k]][[1]]$n
}


### save
save(R2_results_clin,
     table_clin,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_clinical_factors.RData"))

```

## Check proportional hazards for exposome + PRS + clinicl risk factor models

```{r ph_check_clin}
load(paste0(path, "/results/incident disease/R2_models_nov_09_2022_clinical_factors.RData"))

sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)
cph_identify_prev <- identify_prev_cases_list[sig_exposures]
surv_list <- surv_list[sig_exposures]
model_exposures <- lapply(exposures_list, paste, collapse = " + ")
model_exposures <- model_exposures[sig_exposures]

j <- 1

test <- as.list(seq_along(cph_identify_prev))
for (k in 1:length(cph_identify_prev)) {
    data <- lapply(all_data_ncd, cph_identify_prev[[k]])
    test[[k]] <- cox.zph(R2_results_clin[[k]][[j]])
    rm(data)
    cat("\r", paste("completed:", paste(k, length(R2_results_clin), sep = "/")))
}

names(test) <- disease_list[sig_exposures]

# save
save(test,
     file = paste0(path, "/results/incident disease/cph_check_nov_09_2022_clinical.RData"))

# return to full surv_list
surv_list <- surv_list_copy

```

## Check residuals for exposome + PRS + clinicl risk factor models

```{r ph_residuals_clin}

# load(paste0(path, "/results/incident disease/cph_check_sept_01_2022_prs.RData"))

### find variables with low p-value from coxzph test
cph_var <- sapply(test, function(x){rownames(as.data.frame(x$table))[which(as.data.frame(x$table)$p < 0.00001)]})
cph_var <- lapply(cph_var, function(x) {x <- x[x != "GLOBAL"]})


### check schoenfield residuals for each

# residuals **could be skewed for age - test further
for (k in 1:length(cph_var$`All-cause mortality`)) {
plot(test$`All-cause mortality`, var = cph_var$`All-cause mortality`[k], resid = TRUE)
}

# residuals look fine
for (k in 1:length(cph_var$`Prostate cancer`)) {
plot(test$`Prostate cancer`, var = cph_var$`Prostate cancer`[k], resid = TRUE)
}

# residuals look fine
plot(test$`Type II diabetes`, var = cph_var$`Type II diabetes`, resid = TRUE)

# residuals look fine
for (k in 1:length(cph_var$`Ischemic heart disease`)) {
plot(test$`Ischemic heart disease`, var = cph_var$`Ischemic heart disease`[k], resid = TRUE)
}

# residuals look fine
for (k in 1:length(cph_var$`Emphysema, COPD`)) {
plot(test$`Emphysema, COPD`, var = cph_var$`Emphysema, COPD`[k], resid = TRUE)
}

# residuals look fine
for (k in 1:length(cph_var$`Chronic liver diseases`)) {
plot(test$`Chronic liver diseases`, var = cph_var$`Chronic liver diseases`[k], resid = TRUE)
}

# residuals **could be skewed for baseline_cvd
for (k in 1:length(cph_var$`Chronic kidney diseases`)) {
plot(test$`Chronic kidney diseases`, var = cph_var$`Chronic kidney diseases`[k], resid = TRUE)
}

# residuals look fine
plot(test$`Rheumatoid arthritis`, var = cph_var$`Rheumatoid arthritis`, resid = TRUE)

# residuals **could be skewed for BMI
plot(test$Osteoarthritis, var = cph_var$Osteoarthritis, resid = TRUE)

# two variables to correct:
## all-cause mortality - age_scaled

```

## Re-run exposome + PRS + clinicl risk factor models with time interaction

```{r ph_run_models_clin}

indicator <- c(
    Cs(
        ACM_event_indicator,
        colon_cancer_event,
        lung_cancer_event,
        eso_cancer_event,
        liver_cancer_event,
        pancreatic_cancer_event,
        brain_cancer_event,
        leukemia_event,
        lymphoma_event,
        breast_cancer_event,
        ovarian_cancer_event,
        prostate_cancer_event,
        diabetes_event,
        CVD_event,
        cerebro_event,
        COPD_event,
        liver_event,
        kidney_event,
        all_dementia_event,
        vasc_dementia_event,
        alzheimers_event,
        parkinsons_event,
        rheumatoid_event,
        macular_event,
        osteoporosis_event,
        osteoarthritis_event
    )
)

surv_time <- c(
    Cs(
        ACM_survival_time,
        colon_cancer_survival_time,
        lung_cancer_survival_time,
        eso_cancer_survival_time,
        liver_cancer_survival_time,
        pancreatic_cancer_survival_time,
        brain_cancer_survival_time,
        leukemia_survival_time,
        lymphoma_survival_time,
        breast_cancer_survival_time,
        ovarian_cancer_survival_time,
        prostate_cancer_survival_time,
        diabetes_survival_time,
        CVD_survival_time,
        cerebro_survival_time,
        COPD_survival_time,
        liver_survival_time,
        kidney_survival_time,
        all_dementia_survival_time,
        vasc_dementia_survival_time,
        alzheimers_survival_time,
        parkinsons_survival_time,
        rheumatoid_survival_time,
        macular_survival_time,
        osteoporosis_survival_time,
        osteoarthritis_survival_time
    )
)

# get list of outcomes to redo with time interaction
redo <- which(names(R2_results_clin) %in% c("All-cause mortality"))

# make index of diseases with at least 1 significant exposure
sig_exposures <- sapply(exposures_list, length)
sig_exposures <- which(sig_exposures != 0)
cph_exposures_list <- clinical_vars[sig_exposures]
cph_identify_prev <- identify_prev_cases_list[sig_exposures]
surv_list_time <- surv_list_tt[sig_exposures]
cph_indicator <- indicator[sig_exposures]
cph_surv <- surv_time[sig_exposures]
ph.test <- as.list(seq(1, length(R2_results_clin)))

# make model exposures
time_int_var <- as.list(seq(1, length(R2_results_clin)))
names(time_int_var) <- disease_list[sig_exposures]

# subset to vars with bad residuals
time_int_var$`All-cause mortality` <- c("age_scaled")

for (k in redo) {
    time_int_var[[k]] <- paste0(time_int_var[[k]], ":Start_time")
}

for (k in seq_along(time_int_var)[-redo]) {
    time_int_var[[k]] <- vector()
}

time_int_var <- lapply(time_int_var, function(x) paste(x, collapse = " + "))
model_exposures <- lapply(cph_exposures_list, paste, collapse = " + ")


library(magrittr)
library(Greg)

# copy old R2 results
R2_results_clin_cph <- R2_results_clin

# redo interaction models
for (k in redo) {
  
    # subset data
    data <- all_data_ncd
    data <- lapply(data, cph_identify_prev[[k]])
    
    # timesplit function
    data <- lapply(data, function(x){
        # reduce data to just colnames of model variables to improve runtime with timesplit
        x <- x[c(cph_exposures_list[[k]], cph_indicator[k], cph_surv[k], "recruitment_age", "age_scaled", "sex")]
        # turn survival time to years to be on same scale as age
        x$time <- as.numeric(x[[cph_surv[k]]] / 365.25)
        # timesplit data
        x <- x %>% 
            timeSplitter(by = 2,  
                         event_var = cph_indicator[k],
                         event_start_status = 0,
                         time_var = "time",
                         time_related_vars = "recruitment_age")
        return(x)
    })
    
    # make sub-list for model results
    R2_results_clin_cph[[k]] <- as.list(seq(1, length(data)))
    ph.test[[k]] <- as.list(seq(1, length(data)))
    # run model in each imputed dataset
    for (j in seq_along(data)) {

        R2_results_clin_cph[[k]][[j]] <-
            coxph(as.formula(
                paste(
                    paste0(
                        surv_list_time[[k]],
                        model_exposures[[k]]),
                    time_int_var[[k]], sep = " + ")
                ), data = data[[j]])
        
        # test ph assumption, stop if not fixed
        ph.test[[k]][[j]] <- cox.zph(R2_results_clin_cph[[k]][[j]])
        # trial <- any(as.data.frame(ph.test$table)$p < 0.00001)
        # if (trial == TRUE) {
            # print(ph.test)
            # stop("non-proportional hazards not fixed")
        # }
    }
    
    # remove
    rm(data)
    invisible(gc())
    
    cat("\r", paste("completed:", paste(k, length(R2_results_clin_cph), sep = "/")))
}

# create table_clin_cph for results
table_clin_cph <- data.frame(matrix(ncol = 5, nrow = length(R2_results_clin_cph)))
colnames(table_clin_cph) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")

# get R2 and C-index for all models
for (k in seq_along(R2_results_clin_cph)) {

    # get r-squared across imputations
    avg_rsq <- c(coxr2(R2_results_clin_cph[[k]][[1]])$rsq,
                 coxr2(R2_results_clin_cph[[k]][[2]])$rsq,
                 coxr2(R2_results_clin_cph[[k]][[3]])$rsq,
                 coxr2(R2_results_clin_cph[[k]][[4]])$rsq,
                 coxr2(R2_results_clin_cph[[k]][[5]])$rsq)
    
    # transform to correlation scale (sqrt)
    avg_rsq <- sapply(avg_rsq, sqrt)
    
    # transform using Fischer's Z
    avg_rsq <- sapply(avg_rsq, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_rsq <- mean(avg_rsq)
    
    # transform coefficients back to r scale
    avg_rsq <- FisherZInv(avg_rsq)
    
    # transform r squared scale
    avg_rsq <- (avg_rsq)^2
    
    # add to table_clin_cph
    table_clin_cph[k,3] <- avg_rsq
    
    
    # get c-index across imputations
    avg_index <- c(summary(R2_results_clin_cph[[k]][[1]])$concordance[1],
                   summary(R2_results_clin_cph[[k]][[2]])$concordance[1],
                   summary(R2_results_clin_cph[[k]][[3]])$concordance[1],
                   summary(R2_results_clin_cph[[k]][[4]])$concordance[1],
                   summary(R2_results_clin_cph[[k]][[5]])$concordance[1])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table_clin_cph
    table_clin_cph[k,2] <- avg_index


}

table_clin_cph$Disease <- names(R2_results_clin_cph)

# add in number of cases and sample sizes
for (k in seq_along(R2_results_clin_cph)[-redo]) {

    table_clin_cph[k,4] <- R2_results_clin_cph[[k]][[1]]$nevent
    table_clin_cph[k,5] <- R2_results_clin_cph[[k]][[1]]$n
}

# use old table for new time split models, since sample will be in millions
for (k in seq_along(R2_results_clin_cph)[redo]) {

    table_clin_cph[k,4] <- table_exp[k,4]
    table_clin_cph[k,5] <- table_exp[k,5]
}


### save
save(R2_results_clin_cph,
     ph.test,
     table_clin_cph,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_clinical_factors_cph.RData"))

```

## Validate exposome + PRS + clinical risk factor model in Scotland/Wales dataset 

```{r prediction_clinical, eval=TRUE}

# subset surv list
indicator_val <- indicator[sig_exposures]
surv_time_val <- surv_time[sig_exposures]

# subset prev cases function list
identify_prev_cases_list_val <- identify_prev_cases_list[sig_exposures]

# get prediction values for each disease
pred_list <- as.list(seq_along(R2_results_clin))
cindex_list_clin <- as.list(seq_along(R2_results_clin))
rsq_list_clin <- as.list(seq_along(R2_results_clin))

table_scotwales_clin <- data.frame(matrix(ncol = 5, nrow = length(cindex_list_clin)))
colnames(table_scotwales_clin) <- c("Disease", "C-index", "R-squared", "Cases", "Sample size")


for (k in seq_along(pred_list)) {
    
    # subset data
    data <- scotwales_ncd
    data <- lapply(data, identify_prev_cases_list_val[[k]])
    
    # make sub-list for model results
    pred_list[[k]] <- as.list(seq(1, length(data)))
    cindex_list_clin[[k]] <- as.list(seq(1, length(data)))
    rsq_list_clin[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        # get linear predictors using scotwales data
        pred_list[[k]][[j]] <-
            predict(R2_results_clin[[k]][[j]],
                    newdata = data[[j]])
        
        # Surv object in scotwales
        yrep <- Surv(data[[j]][[surv_time_val[k]]], data[[j]][[indicator_val[k]]])
        # calculate concordance outcomes using linear predictors and Surv object
        rs.corr <- rcorr.cens(x = pred_list[[k]][[j]], S = yrep)
        # take Dxy, negate, and then get C using formula Dxy=2*(C0.5)
        cindex_list_clin[[k]][[j]] <- (-rs.corr[2]/2) + 0.5
        
        # Get R-squared using predicted values from scotwales data
        cox <- coxph(yrep ~ pred_list[[k]][[j]], data = data[[j]])
        rsq_list_clin[[k]][[j]] <- coxr2(cox)
        
    }
    
    table_scotwales_clin[k,4] <- as.numeric(table(data[[1]][indicator_val[k]])[2])
    table_scotwales_clin[k,5] <- nrow(data[[1]])
    
    # record sample size
    rm(data)
    
    cat("\r", paste("completed:", paste(k, length(pred_list), sep = "/")))
    
}

### For further discussion of Dxy calculation, see comment here from Harrell on using linear predictors with rcorr.cens: https://stats.stackexchange.com/questions/48298/computing-c-index-for-an-external-validation-of-a-cox-ph-model-with-r#comment93976_48369


# get C-index for all models
for (k in seq_along(cindex_list_clin)) {
    
    # Rsq
    avg_index <- c(rsq_list_clin[[k]][[1]]$rsq,
                   rsq_list_clin[[k]][[2]]$rsq,
                   rsq_list_clin[[k]][[3]]$rsq,
                   rsq_list_clin[[k]][[4]]$rsq,
                   rsq_list_clin[[k]][[5]]$rsq)
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_clin[k,3] <- avg_index
    
    # C-index
    avg_index <- c(cindex_list_clin[[k]][[1]],
                   cindex_list_clin[[k]][[2]],
                   cindex_list_clin[[k]][[3]],
                   cindex_list_clin[[k]][[4]],
                   cindex_list_clin[[k]][[5]])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_clin[k,2] <- avg_index

}

table_scotwales_clin$Disease <- names(R2_results_clin)

### save
save(rsq_list_clin,
     cindex_list_clin,
     table_scotwales_clin,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_clinical.RData"))

```

## Validate exposome + PRS + clinical risk factor model in Scotland/Wales dataset - adding time interaction results

```{r prediction_clin_cph, eval=TRUE}

# get index of outcomes to redo with time interaction
redo <- which(names(R2_results_clin_cph) %in% c("All-cause mortality"))

# subset surv list
indicator_val <- indicator[sig_exposures]
surv_time_val <- surv_time[sig_exposures]

# subset prev cases function list
identify_prev_cases_list_val <- identify_prev_cases_list[sig_exposures]
cph_exposures_list <- clinical_vars[sig_exposures]
cph_indicator <- indicator[sig_exposures]

# list for prediction values for each disease
pred_list <- as.list(seq_along(R2_results_clin_cph))
cindex_list_clin_cph <- cindex_list_clin
rsq_list_clin_cph <- rsq_list_clin

table_scotwales_clin_cph <- table_scotwales_clin


for (k in redo) {
    
    # subset data
    data <- scotwales_ncd
    data <- lapply(data, identify_prev_cases_list_val[[k]])
    data2 <- data
    
    # timesplit function
    data <- lapply(data, function(x){
        # reduce data to just colnames of model variables to improve runtime with timesplit
        x <- x[c(cph_exposures_list[[k]], cph_indicator[k], cph_surv[k], "recruitment_age", "age_scaled", "sex")]
        # turn survival time to years to be on same scale as age
        x$time <- as.numeric(x[[cph_surv[k]]] / 365.25)
        # timesplit data
        x <- x %>% 
            timeSplitter(by = 2,  
                         event_var = cph_indicator[k],
                         event_start_status = 0,
                         time_var = "time",
                         time_related_vars = "recruitment_age")
            return(x)
        })
    
    # make sub-list for model results
    pred_list[[k]] <- as.list(seq(1, length(data)))
    cindex_list_clin_cph[[k]] <- as.list(seq(1, length(data)))
    rsq_list_clin_cph[[k]] <- as.list(seq(1, length(data)))
    
    # run model in each imputed dataset
    for (j in seq_along(data)) {
        
        # get linear predictors using scotwales data
        pred_list[[k]][[j]] <-
            predict(R2_results_clin_cph[[k]][[j]],
                    newdata = data[[j]])
    
            # Surv object in scotwales using interval time for cph interaction
            yrep <- Surv(data[[j]][["Start_time"]], data[[j]][["Stop_time"]], data[[j]][[indicator_val[k]]])
            # Get R-squared using predicted values from scotwales data
            cox <- coxph(yrep ~ pred_list[[k]][[j]], data = data[[j]])
            cindex_list_clin_cph[[k]][[j]] <- summary(cox)$concordance[1]
            rsq_list_clin_cph[[k]][[j]] <- coxr2(cox)
    }
    
    
    table_scotwales_clin_cph[k,4] <- as.numeric(table(data2[[1]][indicator_val[k]])[2])
    table_scotwales_clin_cph[k,5] <- nrow(data2[[1]])
    
    # record sample size
    rm(data, data2)
    
    cat("\r", paste("completed:", paste(k, length(pred_list), sep = "/")))
    
}

### For further discussion of Dxy calculation, see comment here from Harrell on using linear predictors with rcorr.cens: https://stats.stackexchange.com/questions/48298/computing-c-index-for-an-external-validation-of-a-cox-ph-model-with-r#comment93976_48369


# get C-index for all models
for (k in seq_along(cindex_list_clin_cph)) {
    
    # Rsq
    avg_index <- c(rsq_list_clin_cph[[k]][[1]]$rsq,
                   rsq_list_clin_cph[[k]][[2]]$rsq,
                   rsq_list_clin_cph[[k]][[3]]$rsq,
                   rsq_list_clin_cph[[k]][[4]]$rsq,
                   rsq_list_clin_cph[[k]][[5]]$rsq)
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_clin_cph[k,3] <- avg_index
    
    # C-index
    avg_index <- c(cindex_list_clin_cph[[k]][[1]],
                   cindex_list_clin_cph[[k]][[2]],
                   cindex_list_clin_cph[[k]][[3]],
                   cindex_list_clin_cph[[k]][[4]],
                   cindex_list_clin_cph[[k]][[5]])
    
    # transform to correlation scale (sqrt)
    avg_index <- sapply(avg_index, sqrt)
    
    # transform using Fischer's Z
    avg_index <- sapply(avg_index, FisherZ)
    
    # pool coefficients (which is just the average)
    avg_index <- mean(avg_index)
    
    # transform coefficients back to r scale
    avg_index <- FisherZInv(avg_index)
    
    # transform r squared scale
    avg_index <- (avg_index)^2
    
    # add to table
    table_scotwales_clin_cph[k,2] <- avg_index

}


### save
save(rsq_list_clin_cph,
     cindex_list_clin_cph,
     table_scotwales_clin_cph,
     file = paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_clin_cph.RData"))

```

## Create combined table of all R^2 and C-index results - England participants

```{r combined_table, eval=FALSE}

load(paste0(path, "/results/incident disease/R2_models_aug_25_2022_age_sex.RData"))
load(paste0(path, "/results/incident disease/R2_models_nov_09_2022_exposome_cph.RData"))
load(paste0(path, "/results/incident disease/R2_models_nov_09_2022_PRS_cph.RData"))
load(paste0(path, "/results/incident disease/R2_models_nov_09_2022_clinical_factors_cph.RData"))


colnames(table_as) <- paste(colnames(table_as), "_ageSex") 
colnames(table_as)[1] <- "Disease"

colnames(table_exp_cph) <- paste(colnames(table_exp_cph), "_exp")
colnames(table_exp_cph)[1] <- "Disease"

colnames(table_prs_cph) <- paste(colnames(table_prs_cph), "_PRS")
colnames(table_prs_cph)[1] <- "Disease"

colnames(table_clin_cph) <- paste(colnames(table_clin_cph), "_clinical") 
colnames(table_clin_cph)[1] <- "Disease"

table_combined <-
    merge(table_as,
          table_exp_cph,
          by = "Disease",
          all.x = TRUE,
          all.y = FALSE)

table_combined <-
    merge(table_combined,
          table_prs_cph,
          by = "Disease",
          all.x = TRUE,
          all.y = FALSE)

table_combined <-
    merge(table_combined,
          table_clin_cph,
          by = "Disease",
          all.x = TRUE,
          all.y = FALSE)

new_order <- colnames(table_combined)[c(1,2,6,10,14,3,7,11,15,4,8,12,16,5,9,13,17)]
row_order <- order(table_combined$`C-index _exp`, decreasing = TRUE)
table_combined <- table_combined[row_order, new_order]

# make combined exposome + PRS table
colnames(table_combined) <- c("Disease", 
                              "C-index age + sex",
                              "C-index exposome", 
                              "C-index exposome + PRS", 
                              "C-index exposome + PRS + clinical",
                              "R2 age + sex", 
                              "R2 exposome", 
                              "R2 exposome + PRS",
                              "R2 exposome + PRS + clinical",
                              "Cases",
                              "Cases exposome",
                              "Cases exposome + PRS",
                              "Cases exposome + PRS + clinical",
                              "Sample",
                              "Sample exposome",
                              "Sample exposome + PRS",
                              "Sample exposome + PRS + clinical")

table_combined_red <- table_combined
table_combined_red$Cases <- paste(format(table_combined$`Cases exposome + PRS + clinical`, big.mark = ","), format(table_combined$Cases, big.mark = ","), sep = " - ")
table_combined_red$Sample <- paste(format(table_combined$`Sample exposome + PRS + clinical`, big.mark = ","), format(table_combined$Sample, big.mark = ","), sep = " - ")

table_combined_red <- table_combined_red[c(1:10,14)]

library(flextable)
table_combinedfl <- 
    flextable(table_combined_red) %>%
    add_header_row(top = TRUE, values = "Table 2. Multi-omics model performance in English UK Biobank participants", colwidths = 11)

table_combinedfl <- 
    colformat_double(x = table_combinedfl,
                     j = c(1:9),
                     digits = 4)

table_combinedfl <- 
    colformat_double(x = table_combinedfl,
                     j = c(10:11),
                     big.mark = ",",
                     digits = 0)


library(officer)
sect_properties <- prop_section(
  page_size = page_size(orient = "landscape"),
  type = "continuous",
  page_margins = page_mar()
)

# save
save_as_docx(table_combinedfl, pr_section = sect_properties, path = paste0(path, "/output/tables/rsq_all_England_nov_09_2022_cph.docx"))

```

## Create combined table of all R^2 and C-index results - Validation datset

```{r combined_table_validation}

load(paste0(path, "/results/incident disease/R2_models_aug_25_2022_scotwales_age_sex.RData"))
load(paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_exposome_cph.RData"))
load(paste0(path, "/results/incident disease/R2_models_nov_09_2022_scotwales_PRS_cph.RData"))
load(paste0(path, "/results/incident disease/R2_models_aug_28_2022_scotwales_clinical_factors.RData"))

### validation
colnames(table_scotwales_as) <- paste(colnames(table_scotwales_as), "_ageSex") 
colnames(table_scotwales_as)[1] <- "Disease"

colnames(table_scotwales_exp_cph) <- paste(colnames(table_scotwales_exp_cph), "_exp")
colnames(table_scotwales_exp_cph)[1] <- "Disease"

colnames(table_scotwales_prs_cph) <- paste(colnames(table_scotwales_prs_cph), "_PRS")
colnames(table_scotwales_prs_cph)[1] <- "Disease"

colnames(table_scotwales_clin_cph) <- paste(colnames(table_scotwales_clin_cph), "_clin")
colnames(table_scotwales_clin_cph)[1] <- "Disease"

table_combined_sw <-
    merge(table_scotwales_as,
          table_scotwales_exp_cph,
          by = "Disease",
          all.x = TRUE,
          all.y = FALSE)

table_combined_sw <-
    merge(table_combined_sw,
          table_scotwales_prs_cph,
          by = "Disease",
          all.x = TRUE,
          all.y = FALSE)

table_combined_sw <-
    merge(table_combined_sw,
          table_scotwales_clin_cph,
          by = "Disease",
          all.x = TRUE,
          all.y = FALSE)

new_order <- colnames(table_combined_sw)[c(1,2,6,10,14,3,7,11,15,4,8,12,16,5,9,13,17)]
row_order <- order(table_combined_sw$`C-index _exp`, decreasing = TRUE)
table_combined_sw <- table_combined_sw[row_order, new_order]

# make combined exposome + PRS table
colnames(table_combined_sw) <- c("Disease", 
                                 "C-index age + sex",
                                 "C-index exposome", 
                                 "C-index exposome + PRS", 
                                 "C-index exposome + PRS + clinical",
                                 "R2 age + sex", 
                                 "R2 exposome", 
                                 "R2 exposome + PRS",
                                 "R2 exposome + PRS + clinical",
                                 "Cases",
                                 "Cases exposome",
                                 "Cases exposome + PRS",
                                 "Cases exposome + PRS + clinical",
                                 "Sample",
                                 "Sample exposome",
                                 "Sample exposome + PRS",
                                 "Sample exposome + PRS + clinical")

table_combined_sw_red <- table_combined_sw[c(1:10,14)]

library(flextable)
table_combined_swfl <- 
    flextable(table_combined_sw_red) %>%
    add_header_row(top = TRUE, values = "Table 3. Multi-omics model performance in Scottish/Welsh UK Biobank participants", colwidths = 11)

table_combined_swfl <- 
    colformat_double(x = table_combined_swfl,
                     j = c(1:9),
                     digits = 4)

table_combined_swfl <- 
    colformat_double(x = table_combined_swfl,
                     j = c(10:11),
                     big.mark = ",",
                     digits = 0)

library(officer)
sect_properties <- prop_section(
  page_size = page_size(orient = "landscape"),
  type = "continuous",
  page_margins = page_mar()
)

# save
save_as_docx(table_combined_swfl, pr_section = sect_properties, path = paste0(path, "/output/tables/rsq_all_scotwales_nov_09_2022_cph.docx"))

```

## Run ANOVA using exposome + PRS + clinical risk factor models

```{r chisq_prep_clinical}

### clinical model

data <- all_data_ncd[[1]]
data2 <- scotwales_ncd[[1]]

# list of mortality + outcomes with c-index > 0.80
disease_index <- c(
    "All-cause mortality",
    "Lung cancer",
    "Emphysema, COPD",
    "All-cause dementia",
    "Vascular dementia",
    "Alzheimer's disease",
    "Type II diabetes",
    "Osteoporosis",
    "Chronic liver diseases",
    "Chronic kidney diseases"
)
    
# index of these diseases
cdis <- which(disease_list %in% disease_index)

library(miceadds)
library(DescTools)

# set necessary options for rms
options(contrasts=c("contr.treatment", "contr.treatment"))

# function to extra variable proportion of chisq explained
anova_df <- function(x) {

    for (k in cdis) {
        
        model_list <- as.list(seq_along(x))
        
        for (i in seq_along(x)){
            model_exposures <- paste(clinical_vars[[k]], collapse = " + ")
            data <- identify_prev_cases_list[[k]](x[[i]])
            S <- Surv(data[[surv_time[k]]], data[[indicator[k]]])
            model_list[[i]] <- cph(as.formula(
                paste("S ~ age_scaled + sex + ",
                      model_exposures)),
                x = TRUE, y = TRUE, data = data)
            
            # df of anova results per variable
            df <- as.data.frame(anova(model_list[[i]]))
            # df <- df[c(1:(nrow(df)-1)), ]
            df$Variable <- rownames(df)
            
            if (i == 1) {
                dw <- df[c(4,2,1)]
            } else {
                dw <- merge(
                    dw, 
                    df[c(1,4)],
                    by = "Variable"
                )
            }
        }
        
        if (k == 1) {

            csub <- which(grepl("Chi-Square", colnames(dw)))
            totals <- unlist(dw[which(dw$Variable == "TOTAL"), csub])
            
            # get percentage of total variation
            for (j in 1:nrow(dw)) {
                dw$pctexp[j] <- sum(dw[j,csub])/sum(totals)
            }
            # add group for disease
            dw$group <- disease_list[k]
            # duplicate
            rw <- dw
            
            # make df of anova results by category
            df2 <- data.frame(matrix(nrow = 4, ncol = 2))
            colnames(df2) <- c("Category", "pctexp")
            df2$Category <- c("age, sex", "exposome", "PRS", "clinical factors")
            
            # get % explained for age + sex
            df2$pctexp[which(df2$Category == "age, sex")] <- 
                sum(dw[which(dw$Variable %in% c("age_scaled", "sex")), csub])/sum(totals)
            
            # get % explained for exposome
            df2$pctexp[which(df2$Category == "exposome")] <- 
                sum(dw[which(dw$Variable %in% exposures_list[[k]]), csub])/sum(totals)
            
            # get % explained for PRS
            df2$pctexp[which(df2$Category == "PRS")] <- 
                sum(dw[which(dw$Variable %in% PRS_list), csub])/sum(totals)
            
            # get % explained for clinical
            clin_only <- clinical_vars[[k]][clinical_vars[[k]] %nin% c(PRS_list, exposures_list[[k]])]
            df2$pctexp[which(df2$Category == "clinical factors")] <- sum(dw[which(dw$Variable %in% clin_only), csub])/sum(totals)
            
            df2$group <- disease_list[k]
            df2$Category <- factor(df2$Category, 
                                   levels = c("PRS",
                                              "exposome", 
                                              "clinical factors", 
                                              "age, sex"))
            
            # duplicate
            rw2 <- df2
            
        } else {
            
            csub <- which(grepl("Chi-Square", colnames(dw)))
            totals <- unlist(dw[which(dw$Variable == "TOTAL"), csub])
            
            # get percentage of total variation
            for (j in 1:nrow(dw)) {
                dw$pctexp[j] <- sum(dw[j,csub])/sum(totals)
            }
            # add group for disease
            dw$group <- disease_list[k]
            # rbind
            rw <- rbind(rw, dw)
            
            # make df of anova results by category
            df2 <- data.frame(matrix(nrow = 4, ncol = 2))
            colnames(df2) <- c("Category", "pctexp")
            df2$Category <- c("age, sex", "exposome", "PRS", "clinical factors")
            
            # get % explained for age + sex
            df2$pctexp[which(df2$Category == "age, sex")] <- 
                sum(dw[which(dw$Variable %in% c("age_scaled", "sex")), csub])/sum(totals)
            
            # get % explained for exposome
            df2$pctexp[which(df2$Category == "exposome")] <- 
                sum(dw[which(dw$Variable %in% exposures_list[[k]]), csub])/sum(totals)
            
            # get % explained for PRS
            df2$pctexp[which(df2$Category == "PRS")] <- 
                sum(dw[which(dw$Variable %in% PRS_list), csub])/sum(totals)
            
            # get % explained for clinical
            clin_only <- clinical_vars[[k]][clinical_vars[[k]] %nin% c(PRS_list, exposures_list[[k]])]
            df2$pctexp[which(df2$Category == "clinical factors")] <- sum(dw[which(dw$Variable %in% clin_only), csub])/sum(totals)
            
            df2$group <- disease_list[k]
            df2$Category <- factor(df2$Category, 
                                   levels = c("PRS",
                                              "exposome", 
                                              "clinical factors", 
                                              "age, sex"))
            # rbind
            rw2 <- rbind(rw2, df2)
        }
        cat("\r", paste("completed:", paste(which(cdis == k), length(cdis), sep = "/")))
    }
    
    # list of two df results
    rw_list <- list(
        "var" = rw,
        "cat" = rw2
    )
    
    return(rw_list)
}


# run model
plots <- anova_df(all_data_ncd)

# save
save(plots,
     file = paste0(path, "/results/incident disease/chisq_models_nov_09_2022_clinical.RData"))

# set as factor
plots$cat$group <- factor(plots$cat$group, levels = disease_index)

# add labels
plots$cat$Category <- as.character(plots$cat$Category)
plots$cat$Category[seq(1, 40, 4)] <- "Age and sex"
plots$cat$Category[seq(2, 40, 4)] <- "Exposome"
plots$cat$Category[seq(3, 40, 4)] <- "PRS"
plots$cat$Category[seq(4, 40, 4)] <- "Clinical factors"
plots$cat$Category <- factor(plots$cat$Category, levels = c(
    "PRS",
    "Exposome",
    "Clinical factors",
    "Age and sex"
))

```

## Run ANOVA using exposome + PRS models

```{r chisq_prep_prs}
### exposome + PRS model

# list of mortality + outcomes with c-index > 0.80
disease_index <- c(
    "All-cause mortality",
    "Lung cancer",
    "Emphysema, COPD",
    "All-cause dementia",
    "Vascular dementia",
    "Alzheimer's disease",
    "Chronic liver diseases"
)
    

PRS_index <- which(disease_list %in% PRS_disease_list)

# index of these diseases
cdis <- which(disease_list %in% disease_index)
comb_vars <- exposures_list
comb_vars[PRS_index] <- PRS_exposures

anova_df_prs <- function(x) {

    for (k in cdis) {
        
        model_list <- as.list(seq_along(x))
        
        for (i in seq_along(x)){
            model_exposures <- paste(comb_vars[[k]], collapse = " + ")
            data <- identify_prev_cases_list[[k]](x[[i]])
            S <- Surv(data[[surv_time[k]]], data[[indicator[k]]])
            model_list[[i]] <- cph(as.formula(paste("S ~ age_scaled + sex + ",
                                                    model_exposures)),
                                   x = TRUE, y = TRUE, data = data)

            # df of anova results per variable
            df <- as.data.frame(anova(model_list[[i]]))
            # df <- df[c(1:(nrow(df)-1)), ]
            df$Variable <- rownames(df)
            
            if (i == 1) {
                dw <- df[c(4,2,1)]
            } else {
                dw <- merge(
                    dw, 
                    df[c(1,4)],
                    by = "Variable"
                )
            }
        }
        
        if (k == 1) {

            csub <- which(grepl("Chi-Square", colnames(dw)))
            totals <- unlist(dw[which(dw$Variable == "TOTAL"), csub])
            
            # get percentage of total variation
            for (j in 1:nrow(dw)) {
                dw$pctexp[j] <- sum(dw[j,csub])/sum(totals)
            }
            # add group for disease
            dw$group <- disease_list[k]
            # duplicate
            rw <- dw
            
            # make df of anova results by category
            df2 <- data.frame(matrix(nrow = 3, ncol = 2))
            colnames(df2) <- c("Category", "pctexp")
            df2$Category <- c("age, sex", "exposome", "PRS")
            
            # get % explained for age + sex
            df2$pctexp[which(df2$Category == "age, sex")] <- 
                sum(dw[which(dw$Variable %in% c("age_scaled", "sex")), csub])/sum(totals)
            
            # get % explained for exposome
            df2$pctexp[which(df2$Category == "exposome")] <- 
                sum(dw[which(dw$Variable %in% exposures_list[[k]]), csub])/sum(totals)
            
            # get % explained for PRS
            df2$pctexp[which(df2$Category == "PRS")] <- 
                sum(dw[which(dw$Variable %in% PRS_list), csub])/sum(totals)
            
            df2$group <- disease_list[k]
            df2$Category <- factor(df2$Category, 
                                   levels = c("PRS",
                                              "exposome", 
                                              "age, sex"))
            
            # duplicate
            rw2 <- df2
            
        } else {
            
            csub <- which(grepl("Chi-Square", colnames(dw)))
            totals <- unlist(dw[which(dw$Variable == "TOTAL"), csub])
            
            # get percentage of total variation
            for (j in 1:nrow(dw)) {
                dw$pctexp[j] <- sum(dw[j,csub])/sum(totals)
            }
            # add group for disease
            dw$group <- disease_list[k]
            # rbind
            rw <- rbind(rw, dw)
            
            # make df of anova results by category
            df2 <- data.frame(matrix(nrow = 3, ncol = 2))
            colnames(df2) <- c("Category", "pctexp")
            df2$Category <- c("age, sex", "exposome", "PRS")
            
            # get % explained for age + sex
            df2$pctexp[which(df2$Category == "age, sex")] <- 
                sum(dw[which(dw$Variable %in% c("age_scaled", "sex")), csub])/sum(totals)
            
            # get % explained for exposome
            df2$pctexp[which(df2$Category == "exposome")] <- 
                sum(dw[which(dw$Variable %in% exposures_list[[k]]), csub])/sum(totals)
            
            # get % explained for PRS
            df2$pctexp[which(df2$Category == "PRS")] <- 
                sum(dw[which(dw$Variable %in% PRS_list), csub])/sum(totals)
            
            df2$group <- disease_list[k]
            df2$Category <- factor(df2$Category, 
                                   levels = c("PRS",
                                              "exposome", 
                                              "age, sex"))
            # rbind
            rw2 <- rbind(rw2, df2)
        }
        cat("\r", paste("completed:", paste(which(cdis == k), length(cdis), sep = "/")))
    }
    
    # list of two df results
    rw_list <- list(
        "var" = rw,
        "cat" = rw2
    )
    
    return(rw_list)
}

# run model
plots2 <- anova_df_prs(all_data_ncd)

# save
save(plots2,
     file = paste0(path, "/results/incident disease/chisq_models_nov_09_2022_prs.RData"))

plots2$cat$group <- factor(plots2$cat$group, levels = disease_index)

# add labels
plots2$cat$Category <- as.character(plots2$cat$Category)
plots2$cat$Category[seq(1, 21, 3)] <- "Age and sex"
plots2$cat$Category[seq(2, 21, 3)] <- "Exposome"
plots2$cat$Category[seq(3, 21, 3)] <- "PRS"
plots2$cat$Category <- factor(plots2$cat$Category, levels = c(
    "PRS",
    "Exposome",
    "Age and sex"
))

```

## Calculate partial R^2 across models

```{r partial_r2_clinical}

disease_index <- c(
    "All-cause mortality",
    "Lung cancer",
    "Emphysema, COPD",
    "All-cause dementia",
    "Vascular dementia",
    "Alzheimer's disease",
    "Type II diabetes",
    "Osteoporosis",
    "Chronic liver diseases",
    "Chronic kidney diseases"
)

# function to calculate partial R^2
partialr2 <- function(x) {
    
    bar_mat <- x[c(1,6:9)]
    bar_mat$partial_R2_exp <- 1 - ((1 - bar_mat$`R2 exposome`) / (1 - bar_mat$`R2 age + sex`))
    bar_mat$partial_R2_prs <- 1 - ((1 - bar_mat$`R2 exposome + PRS`) / (1 - bar_mat$`R2 exposome`))
    bar_mat$partial_R2_clin <- 
        ifelse(!is.na(bar_mat$partial_R2_prs),
               1 - ((1 - bar_mat$`R2 exposome + PRS + clinical`) / (1 - bar_mat$`R2 exposome + PRS`)),
               1 - ((1 - bar_mat$`R2 exposome + PRS + clinical`) / (1 - bar_mat$`R2 exposome`))
        )
    
    for (k in seq_along(disease_index)){
        
        if (k == 1) {
            
            mat <- as.data.frame(matrix(nrow = 4, ncol = 3))
            colnames(mat) <- c("Model", "R\u00B2", "Partial R\u00B2")
            mat$Model <- c("Age, sex", "+Exposome", "+PRS", "+Clinical")
            mat$"R\u00B2" <- as.numeric(bar_mat[which(bar_mat$Disease == disease_index[k]), 2:5])
            mat$"Partial R\u00B2" <- c(NA, as.numeric(bar_mat[which(bar_mat$Disease == disease_index[k]), 6:8]))
            
            matw <- mat
            
        } else {
            
            mat <- as.data.frame(matrix(nrow = 4, ncol = 3))
            colnames(mat) <- c("Model", "R\u00B2", "Partial R\u00B2")
            mat$Model <- c("Age, sex", "+Exposome", "+PRS", "+Clinical")
            mat$"R\u00B2" <- as.numeric(bar_mat[which(bar_mat$Disease == disease_index[k]), 2:5])
            mat$"Partial R\u00B2" <- c(NA, as.numeric(bar_mat[which(bar_mat$Disease == disease_index[k]), 6:8]))
            
            matw <- cbind(matw, mat)
        }
    }
    
    mats <- list(
        total = bar_mat,
        partial = matw
    )
    return(mats)

}

# run on table of R^2 results from validation data
partials <- partialr2(table_combined_sw_red)

# create table
ptable <- as.data.frame(partials$partial)
mcols <- which(grepl("Model",names(ptable)))
mcols <- mcols[-1]
ptable <- ptable[-mcols]
headers <- colnames(partials$partial)[-mcols]

for (k in 1:length(colnames(ptable))) {
    headers[k] <- paste0(headers[k], paste0(rep("\r", k), collapse = ""), collapse = "")
}
colnames(ptable) <- headers
nums <- sapply(ptable, is.numeric)
nums <- as.numeric(nums)
nums <- which(nums == 1)
ptable[nums] <- format(round(ptable[nums], digits = 4), nsmall = 4)
for (k in nums) {
    ptable[[k]][which(ptable[[k]] == "    NA")] <- "-"
}

```

## Create Fig. 8 plots - model composition and partial R^2 

```{r combined_plot}

# partial R2 table
pt <- 
    flextable(ptable) %>% 
    add_header_row(top = TRUE, values = c("", disease_index), colwidths = c(1, rep(2, 10)))  %>% 
    add_header_row(top = TRUE, values = "Partial R\u00B2 across outcomes - Scottish/Welsh validation set", colwidths = 21) %>% 
    align(align = "left", part = "all") %>% 
    align(align = "center", part = "header", i = 2) %>% 
    align(align = "center", part = "body", j = seq(2,21))


pt <- vline(
    pt,
    j = c(1, seq(3,20,2)),
    part = "body"
)

ptr <- pt %>%
    as_raster()

gg2 <- ggplot() +
    theme_void() +
    annotation_custom(rasterGrob(ptr), xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf)


# clinical model
g1 <- ggplot(plots$cat, aes(x = group, y = pctexp, fill = Category)) + 
    geom_bar(position = "fill", stat = "identity") +
    scale_fill_viridis(discrete = T, direction = -1) +
    theme_classic() +
    ylab("") +
    xlab("") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
    scale_y_continuous(expand = expansion(mult = c(0.025, 0.05)))


# exposome, PRS model
g2 <- ggplot(plots2$cat, aes(x = group, y = pctexp, fill = Category)) + 
    geom_bar(position = "fill", stat = "identity") +
    # scale_fill_manual(values = viridis(4)[c(4,3,1)]) +
    scale_fill_viridis(discrete = T, direction = -1) +
    theme_classic() +
    ylab("Relative proportion of total \u03C7\u00B2") +
    xlab("") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
    scale_y_continuous(expand = expansion(mult = c(0.025, 0.05)))

g1t <- g1 +
    theme(axis.text.x = element_text(angle = 45, 
                                     hjust = 1, size = 13, face = "bold"),
          axis.text.y = element_text(size = 11),
          axis.title.y = element_text(size = 16, 
                                      margin = margin(r = 20), face = "bold"),
          legend.text = element_text(size = 12),
          legend.title = element_blank(),
          legend.position = "bottom") 
    
g2t <- g2 +
    theme(axis.text.x = element_text(angle = 45, 
                                     hjust = 1, size = 14, face = "bold"),
          axis.text.y = element_text(size = 11),
          axis.title.y = element_text(size = 16, 
                                      margin = margin(r = 20), face = "bold"),
          legend.text = element_text(size = 12),
          legend.title = element_blank(),
          legend.position = "bottom") 

library(cowplot)

# create plot and save
setwd(paste0(path, "/output/incident disease"))
jpeg("chisq_partial_R2_nov_09_2022.jpeg",
    width = 16,
    height = 12,
    units = "in",
    res = 1200)

x <- plot_grid(g2t, g1t, ncol = 2, nrow = 1, rel_widths = c(1,1.35), rel_heights = c(1,1.1), labels = c("", "b"), label_size = 18, align = "v",
               axis = "b")

plot_grid(NULL, x, NULL,
          NULL, gg2, NULL, 
          ncol = 3,
          nrow = 2,
          labels = c("a", "", "", "c", "", ""),
          label_size = 18,
          hjust = -1.5,
          vjust = c(1.5, 1.5, 1.5, 5, 1.5, 1.5),
          axis = "l",
          align = "v",
          rel_heights = c(1.8, 1),
          rel_widths = c(0.1, 1.8, 0.05)
)
dev.off()

```

## Make pretty exposure labels

```{r pretty_labels}

pretty_lables <- function(x) {
    
    x$pretty_labels <- x$Variable
    # make pretty labels for plot
    # this section will need to be tweaked for each new set of results
    x$pretty_labels <-
        gsub("_England", "", 
             x$pretty_labels, 
             fixed = TRUE)
    x$pretty_labels <-
        gsub("_", " ",
             x$pretty_labels,
             fixed = TRUE)
    x$pretty_labels <-
        gsub("hshld", "household", 
             x$pretty_labels, 
             fixed = TRUE)
    x$pretty_labels <-
        gsub("freq", "frequency", 
             x$pretty_labels, 
             fixed = TRUE)
    x$pretty_labels <-
        gsub("prop", "prop.",
             x$pretty_labels,
             fixed = TRUE)
    x$pretty_labels <- 
        gsub("pack", "smoking pack",
             x$pretty_labels,
             fixed = TRUE)
    x$pretty_labels <- 
        gsub("tobacco", "tobacco use",
             x$pretty_labels,
             fixed = TRUE)
    x$pretty_labels <- 
        gsub("environ", "environment",
             x$pretty_labels,
             fixed = TRUE)
    x$pretty_labels <- 
        gsub("10yrs", "10 yrs",
             x$pretty_labels,
             fixed = TRUE)
    
    x$pretty_labels[x$Variable == "worry_embarassment"] <- 
        "worry after embarassment"
    
    x$pretty_labels[x$Variable == "psychiatrist_visit_mental_health"] <- 
        "psychiatrist visit for MH"
    x$pretty_labels[x$Variable == "doctor_visit_mental_health"] <- 
        "doctor visit for MH"
    
    x$pretty_labels[x$Variable == "maternal_smoking"] <- 
        "maternal smoking around birth"
    
    x$pretty_labels[x$Variable == "open_fire_heat"] <- 
        "uses open fire for heating"
    x$pretty_labels[x$Variable == "gas_hob_heat"] <- 
        "uses gas hob for heating"
    x$pretty_labels[x$Variable == "gas_fire_heat"] <- 
        "uses gas fire for heating"
    
    x$pretty_labels[x$Variable == "mobile_phone_duration"] <- 
        "years using mobile phone"
    x$pretty_labels[x$Variable == "mobile_phone_2yrs"] <- 
        "mobile phone use now vs. 2 yrs ago"
    
    x$pretty_labels[x$Variable == "sex_age_first"] <- 
        "age first sexual intercourse"
    
    x$pretty_labels[x$Variable == "easy_wake"] <- 
        "easy to wake in the morning"
    
    x$pretty_labels[x$Variable == "hshld_vehicles"] <- 
        "no. household vehicles"
    
    x$pretty_labels[x$Variable == "hshld_number"] <- 
        "no. people living in household"
    
    x$pretty_labels[x$Variable == "own_or_rent"] <- 
        "home ownership"
    
    x$pretty_labels[x$Variable == "air_pollution_PC1"] <- 
        "air pollution"
    
    x$pretty_labels[x$Variable == "greenspace_PC1"] <- 
        "greenspace"
    
    return(x)
    
}

```

## Create Fig. 7 and supplementary figures - importance per exposure for each disease

```{r exposome_chisq_per_outcome}

# list of mortality + outcomes with c-index > 0.80
disease_index <- c(
    "All-cause mortality",
    "Lung cancer",
    "Emphysema, COPD",
    "All-cause dementia",
    "Vascular dementia",
    "Alzheimer's disease",
    "Type II diabetes",
    "Osteoporosis",
    "Chronic liver diseases",
    "Chronic kidney diseases"
)
    
# index of these diseases
cdis <- which(disease_list %in% disease_index)

# function to extra variable proportion of chisq explained
anova_df_exp <- function(x) {

    df_list <- as.list(rep(1, length(disease_list)))
    
     for (k in cdis) {
        
        model_list <- as.list(seq_along(x))
        
        for (i in seq_along(x)){
            model_exposures <- paste(exposures_list[[k]], collapse = " + ")
            data <- identify_prev_cases_list[[k]](x[[i]])
            S <- Surv(data[[surv_time[k]]], data[[indicator[k]]])
            model_list[[i]] <- cph(as.formula(paste("S ~ age_scaled + sex + ",
                                                    model_exposures)),
                                   x = TRUE, y = TRUE, data = data)

            # df of anova results per variable
            df <- as.data.frame(anova(model_list[[i]]))
            # df <- df[c(1:(nrow(df)-1)), ]
            df$Variable <- rownames(df)
            
            if (i == 1) {
                dw <- df[c(4,2,1)]
            } else {
                dw <- merge(
                    dw, 
                    df[c(1,4)],
                    by = "Variable"
                )
            }
        }
        
        csub <- which(grepl("Chi-Square", colnames(dw)))
        totals <- unlist(dw[which(dw$Variable == "TOTAL"), csub])
        
        # get percentage of total variation
        for (j in 1:nrow(dw)) {
            dw$pctexp[j] <- sum(dw[j,csub])/sum(totals)
        }
        # add group for disease
        dw$group <- disease_list[k]
        # duplicate
        rw <- dw
        
        #rename
        df_list[[k]] <- dw
        
        cat("\r", paste("completed:", 
                        paste(which(cdis == k), length(cdis), sep = "/")))
     }
    names(df_list) <- disease_list
    exclude <- which(sapply(df_list, is.numeric))
    df_list <- df_list[-exclude]
    return(df_list)
}    
    
# run model
plots3 <- anova_df_exp(all_data_ncd)

# save
save(plots3,
     file = paste0(path, "/results/incident disease/chisq_models_nov_09_2022_exposome_indiv.RData"))

load(paste0(path, "/results/incident disease/chisq_models_nov_09_2022_exposome_indiv.RData"))

# make pretty labels
plots3 <- lapply(plots3, pretty_lables)

# formatting
plots3 <- lapply(plots3, function(x){
    
    names(x)[3:7] <- c("chisq1","chisq2","chisq3","chisq4","chisq5")
    x <- x[which(x$pretty_labels != "TOTAL"), ]
    x <- merge(x,
              subset(categories, select = c("Variable", "Category")),
              by = "Variable",
              all.x = TRUE,
              all.y = FALSE)
    x$Category[which(x$Variable == "age_scaled")] <- "Age and sex"
    x$Category[which(x$Variable == "sex")] <- "Age and sex"
    x <- x[order(x$pctexp, decreasing = TRUE), ]
    var_order <- x$pretty_labels
    x$pretty_labels <- factor(x$pretty_labels, levels = var_order)

    return(x)
})


######### exposome plots

## mortality

library(ggforce)
library(ggsci)

load(paste0(path,"/results/color_palette.RData"))
# color_map$color[which(color_map$sequence == "Age and sex")] <- 
    # color_map$color[which(color_map$sequence == "General health")]

mort <- plots3[[disease_index[1]]]

color_map$sequence[which(color_map$sequence == "Medications")] <- "Supplements"
color_map$color[which(color_map$sequence == "Supplements")] <- 
    color_map$color[which(color_map$sequence == "Urine biomarkers")]

pal <- merge(mort, 
             color_map, 
             by.x = "Category",
             by.y = "sequence", 
             all = T)

pal <- pal[which(!is.na(pal$Variable)), ]
pal$color[which(pal$Category == "Age and sex")] <- "#756CD8"

pal <- unique(pal$color)

g3a <- ggplot(mort, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab("") +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1)),
                       limits = c(0,0.2), breaks = seq(0,0.2, by = 0.05)) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01))) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
          axis.title.y = element_text(face = "bold", margin = margin(r = 20)),
          legend.position = "none"
          )

# try scaling y-axis
g3abc <- ggplot(mort, aes(x = pretty_labels, y = sqrt(pctexp), fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2 (square root)") +
    xlab("") +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01))) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
          axis.title.y = element_text(face = "bold", margin = margin(r = 20)),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")
    )
    

g3ab <- ggplot(mort, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("") +
    xlab("") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1)),
                       limits = c(0,0.00832), breaks = seq(0,0.008, by = 0.002)) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)),
                     limits = mort$pretty_labels[mort$pretty_labels %nin% c("age scaled", "sex", "smoking status", "smoking pack years prop.")])
    

# save two-plot version
setwd(paste0(path, "/output/incident disease"))
jpeg("exposome_mortality_chisq_nov_09_2022.jpeg",
    width = 16,
    height = 7,
    units = "in",
    res = 1200)

plot_grid(NULL, g3a, g3ab, nrow = 1, ncol = 3, rel_widths = c(0.03,1,1.3), labels = c("", "a", "b"))

dev.off()

# save single plot version
setwd(paste0(path, "/output/incident disease"))
jpeg("exposome_mortality_chisq_nov_09_2022_nosplit.jpeg",
    width = 10,
    height = 7,
    units = "in",
    res = 1200)

g3abc

dev.off()


## lung cancer
lung <- plots3[[disease_index[2]]]

g3b <- ggplot(lung, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[2], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## COPD
copd <- plots3[[disease_index[2]]]

g3c <- ggplot(copd, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[3], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## all-cause dementia
adem <- plots3[[disease_index[4]]]

g3d <- ggplot(adem, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[4], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## vascular dementia
vdem <- plots3[[disease_index[5]]]

g3e <- ggplot(vdem, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[5], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## AD
AD <- plots3[[disease_index[6]]]

g3f <- ggplot(AD, aes(x = pretty_labels, y = pctexp, fill = Category
)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[6], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## T2D
t2d <- plots3[[disease_index[7]]]

g3g <- ggplot(t2d, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[7], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## osteoporosis
osteo <- plots3[[disease_index[8]]]

g3h <- ggplot(osteo, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[8], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## chronic liver disease
cld <- plots3[[disease_index[9]]]

g3i <- ggplot(cld, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[9], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))

## chronic kidney disease
ckd <- plots3[[disease_index[10]]]

g3j <- ggplot(ckd, aes(x = pretty_labels, y = pctexp, fill = Category)) + 
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = pal) +
    theme_classic() +
    ylab("Proportion of total \u03C7\u00B2") +
    xlab(paste(disease_index[10], "model exposures")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
          legend.text = element_text( size = 10),
          legend.title = element_text( size = 12),
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(0.4, "cm")) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
    scale_x_discrete(expand = expansion(mult = c(0.025, 0.01)))


# save plots
ggsave(g3b, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[2])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3c, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[3])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3d, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[4])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3e, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[5])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3f, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[6])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3g, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[7])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3h, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[8])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3i, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[9])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

ggsave(g3j, filename = paste0(paste0(paste0(path, "/output/incident disease/3. exposome chi-square per disease/exposome_"), gsub(" ", "_", disease_index[10])), "_chisq_nov_09_2022.png"), width = 9, height = 6)

```
